---
title: "Correlations"
author: "Aaron R. Caldwell"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: TRUE
editor_options: 
  chunk_output_type: console
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Correlations}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

The TOSTER package provides several functions for calculating and analyzing correlations. These functions extend beyond traditional correlation tests by offering equivalence testing capabilities and robust correlation methods. The included functions are based on research by @goertzen2010 (`z_cor_test` & `compare_cor`), and @wilcox2011introduction (`boot_cor_test`)^[The bootstrapped functions were adapted from code posted by Rand Wilcox on his website, with modifications inspired by Guillaume Rousselet's `bootcorci` R package, available on GitHub: https://github.com/GRousselet].

# Simple Correlation Test

Basic tests of association can be performed with the `z_cor_test` function. This function is styled after R's built-in `cor.test` function but uses Fisher's z transformation as the basis for all significance tests (p-values). Despite this difference in methodology, the confidence intervals are typically very similar to those produced by `cor.test`.

```{r}
library(TOSTER)
# Base R correlation test
cor.test(mtcars$mpg, mtcars$qsec)

# TOSTER's z-transformed correlation test
z_cor_test(mtcars$mpg, mtcars$qsec)
```

Like `cor.test`, the `z_cor_test` function supports Spearman and Kendall correlation coefficients:

```{r}
# Spearman correlation
z_cor_test(mtcars$mpg,
           mtcars$qsec,
           method = "spear") # Short form accepted; "spearman" also works

# Kendall correlation
z_cor_test(mtcars$mpg,
           mtcars$qsec,
           method = "kendall")
```

# Advantages of z_cor_test

The main advantage of `z_cor_test` over the standard `cor.test` is its ability to perform equivalence testing (TOST) or any hypothesis test where the null hypothesis isn't zero. This makes it particularly useful for research questions focused on demonstrating practical equivalence or testing against specific correlation thresholds. The main disadvantage is that it relies on the Fisher's z transformation, which can be less accurate when assumptions are violated or when outliers are present. In such cases, the bootstrapped methods may provide more reliable results.

```{r}
# Equivalence test with null boundary of 0.4
z_cor_test(mtcars$mpg,
           mtcars$qsec,
           alternative = "e", # e for equivalence
           null = .4)
```

In this example, we're testing whether the correlation is equivalent to zero within the boundaries of ±0.4.

# Using Summary Statistics

A key advantage of TOSTER is the ability to perform correlation tests using only summary statistics, which is particularly useful when reviewing published literature or working with limited data access. The `corsum_test` function enables this functionality:

```{r}
# Testing a correlation of 0.121 from a sample of 105 paired observations
corsum_test(r = .121,
            n = 105,
            alternative = "e",
            null = .4)
```

This example tests whether a correlation of 0.121 from a sample of 105 paired observations is equivalent to zero within the boundaries of ±0.4.

# Bootstrapped Correlation Test

For more robust analyses when raw data is available, TOSTER provides the `boot_cor_test` function. This bootstrapping approach generally produces more reliable results than Fisher's z-based tests, especially when outliers are present or distribution assumptions are violated.

```{r}
set.seed(993) # Setting seed for reproducibility
boot_cor_test(mtcars$mpg,
           mtcars$qsec,
           alternative = "e",
           null = .4)

# Bootstrapped Spearman correlation
boot_cor_test(mtcars$mpg,
           mtcars$qsec,
           method = "spear",
           alternative = "e",
           null = .4)

# Bootstrapped Kendall correlation
boot_cor_test(mtcars$mpg,
           mtcars$qsec,
           method = "ken", # Short form accepted
           alternative = "e",
           null = .4)
```

## Bootstrap Confidence Intervals

The `boot_ci` argument selects the bootstrap confidence interval, and the p-value is always obtained by inverting that same interval, so the two always agree.

- `"auto"` (default): `"stud"` for Pearson's r and `"bca"` for all other correlations. In simulations, the studentized interval kept Pearson's r near nominal coverage for skewed, heteroscedastic, and heavy-tailed data, where BCa under-covered; for the rank and robust correlations BCa did as well or better.
- `"bca"`: bias-corrected and accelerated percentile interval.
- `"perc"`: percentile interval.
- `"basic"`: basic (reflected percentile) interval.
- `"stud"`: studentized (bootstrap-t) interval, available for Pearson, Spearman, and Kendall correlations.

The method that was used is returned in the `boot_ci` element of the result. The Pearson examples above therefore used the studentized interval, and the Spearman and Kendall examples used BCa. Here is a studentized interval for Spearman's rho:

```{r}
set.seed(993)
boot_cor_test(mtcars$mpg,
              mtcars$qsec,
              method = "spearman",
              boot_ci = "stud",
              alternative = "e",
              null = .4)
```

The percentile and BCa intervals give the same answer on any monotone scale. The basic and studentized intervals do not, so the `boot_scale` argument sets the scale on which they are computed: the Fisher z scale (`boot_scale = "z"`, the default), or the correlation scale (`boot_scale = "r"`). The z scale keeps the interval within $[-1, 1]$ and is closer to variance-stabilizing.

### Standard Errors for the Studentized Bootstrap

The studentized bootstrap standardizes every bootstrap replicate by its own standard error, $t^* = (\hat\theta^* - \hat\theta) / \widehat{SE}^*$, and uses the distribution of $t^*$ in place of a normal or $t$ reference distribution. This only improves on the basic interval if $\widehat{SE}^*$ is estimated from the data in each resample. The usual normal-theory standard errors of Fisher's z (e.g., $1/\sqrt{n-3}$ for Pearson's r) depend only on the sample size and so cannot be used for this.

`boot_cor_test()` therefore uses an influence-function (sandwich) standard error, which is computed from the data and does not assume bivariate normality:

$$
\widehat{SE}(\hat\rho) = \frac{1}{n}\sqrt{\sum_{i=1}^{n} \psi_i^2},
$$

where $\psi_i$ is the empirical influence of observation $i$ on the coefficient. On the z scale the delta method gives $\widehat{SE}(\hat z) = \widehat{SE}(\hat\rho) / (1 - \hat\rho^2)$. The influence functions are:

**Pearson's r.** The starting point is the asymptotic distribution-free (fourth-moment) influence function [@steiger1982]. With $z_{x,i}$ and $z_{y,i}$ the standardized observations (using denominator $n$),

$$
\psi^0_i = z_{x,i} z_{y,i} - \frac{r}{2}\left(z_{x,i}^2 + z_{y,i}^2\right).
$$

Under bivariate normality this reduces to the familiar $\text{Var}(r) \approx (1-\rho^2)^2/n$. On its own, though, it gives standard errors that are too small in small samples and with heavy-tailed data, because a sample under-represents the extreme points that dominate the variance of $r$. Following the HC4 heteroscedasticity-consistent estimator for regression [@cribari2004], which @wilcox2011introduction also uses for testing correlations, each influence value is inflated according to its leverage:

$$
\psi_i = \psi^0_i\,(1 - h_i)^{-\delta_i/2}, \qquad h_i = \frac{1}{n} + \frac{D_i^2}{n-1}, \qquad \delta_i = \min\left(4, \frac{n h_i}{3}\right),
$$

where $D_i$ is the Mahalanobis distance of $(x_i, y_i)$ from the bivariate mean (so the $h_i$ sum to 3), and $h_i$ is capped at 0.99. This leverage correction of the ADF influence function is an adaptation made for TOSTER. It was chosen because in simulations (normal, heteroscedastic, $t_5$, $t_3$, and discrete data; $n$ = 30 and 80) it kept the studentized interval's coverage near nominal, including for $t_3$ data, where the uncorrected standard error under-covered badly. It is slightly conservative for strongly heteroscedastic data. When the data have infinite fourth moments, no standard error of $r$ is consistent, and a robust correlation (below) is still the better choice.

**Spearman's rho.** Spearman's rho is Pearson's r computed on the mid-distribution transforms $u_i = (R(x_i) - 1/2)/n$ and $v_i = (R(y_i) - 1/2)/n$, where $R$ denotes midranks. Its influence function adds terms for estimating these transforms to the Pearson influence function of $(u, v)$. With $\omega^x_{ij} = \mathbf{1}(x_j > x_i) + \tfrac{1}{2}\mathbf{1}(x_j = x_i)$ (and $\omega^y_{ij}$ likewise), the influence of observation $i$ on each moment is

$$
\begin{aligned}
d_i(\overline{uv}) &= u_i v_i + \tfrac{1}{n}\textstyle\sum_j \omega^x_{ij} v_j + \tfrac{1}{n}\sum_j \omega^y_{ij} u_j - 3\,\overline{uv}, \\
d_i(\bar u) &= u_i + \tfrac{1}{n}\textstyle\sum_j \omega^x_{ij} - 2\bar u, \\
d_i(\overline{u^2}) &= u_i^2 + \tfrac{2}{n}\textstyle\sum_j \omega^x_{ij} u_j - 3\,\overline{u^2},
\end{aligned}
$$

(similarly for $v$), and the delta method for $r_s = C/\sqrt{S_u S_v}$, with $C = \overline{uv} - \bar u \bar v$ and $S_u = \overline{u^2} - \bar u^2$, gives

$$
\psi_i = \frac{d_i(C)}{\sqrt{S_u S_v}} - \frac{r_s}{2}\left(\frac{d_i(S_u)}{S_u} + \frac{d_i(S_v)}{S_v}\right),
$$

where $d_i(C) = d_i(\overline{uv}) - \bar v\, d_i(\bar u) - \bar u\, d_i(\bar v)$ and $d_i(S_u) = d_i(\overline{u^2}) - 2\bar u\, d_i(\bar u)$. Without ties this is the influence function of Spearman's rho derived by @croux2010. The tie terms keep the standard error accurate for discrete data, and for bootstrap resamples, which always contain ties.

**Kendall's tau.** Kendall's tau-b is a ratio of U-statistics, $\tau_b = \bar a / \sqrt{\bar b^x \bar b^y}$, where

$$
a_i = \frac{1}{n-1}\sum_{j \ne i} \text{sign}(x_i - x_j)\,\text{sign}(y_i - y_j), \qquad b^x_i = \frac{1}{n-1}\sum_{j \ne i} \mathbf{1}(x_i \ne x_j),
$$

and $b^y_i$ is defined likewise. Hoeffding's projection [@hoeffding1948] and the delta method give

$$
\psi_i = 2\left[\frac{a_i - \bar a}{\sqrt{\bar b^x \bar b^y}} - \frac{\tau_b}{2}\left(\frac{b^x_i - \bar b^x}{\bar b^x} + \frac{b^y_i - \bar b^y}{\bar b^y}\right)\right].
$$

Without ties this is the usual U-statistic variance, $\frac{4}{n}\widehat{\text{Var}}(a_i)$. Computing it takes $O(n^2)$ operations per resample, so the studentized Kendall interval is slower in large samples.

Under independence, all three standard errors reduce to the familiar values, $\text{Var}(r) \approx \text{Var}(r_s) \approx 1/n$ and $\text{Var}(\tau) \approx 4/(9n)$. Because each pivot divides by an estimated standard error, studentized intervals can be wide or erratic in small samples (roughly $n < 30$). This matters most for the rank correlations. Kendall's standard error, for example, is nearly unbiased even at $n = 20$, but it varies a lot between samples (especially with ties) and is strongly correlated with the estimate. In simulations at $n = 30$, the studentized Spearman and Kendall intervals under-covered by up to about 2.5 percentage points for heavy-tailed or tied data, while the BCa interval stayed closer to nominal; at $n = 80$ both were accurate. For Spearman's rho and Kendall's tau with fewer than about 50 observations, use `boot_ci = "bca"`, which is what `"auto"` selects for these methods. The studentized interval is not available for the robust correlations below, which lack a closed-form standard error.

## Robust Correlation Methods

The `boot_cor_test` function also provides access to robust correlation methods that are less sensitive to outliers and violations of normality:

```{r}
# Winsorized correlation with 10% trimming
boot_cor_test(mtcars$mpg,
           mtcars$qsec,
           method = "win",
           alternative = "e",
           null = .4,
           tr = .1) # Set trim amount (default is 0.2)

# Percentage bend correlation
boot_cor_test(mtcars$mpg,
           mtcars$qsec,
           method = "bend",
           alternative = "e",
           null = .4,
           beta = .15) # Beta parameter controlling resistance to outliers
```

The Winsorized correlation reduces the impact of outliers by replacing extreme values with less extreme values. The percentage bend correlation is another robust method that downweights the influence of outliers in the calculation.

# Comparing Correlations

TOSTER provides tools for comparing correlations between independent groups or studies. This is useful for testing differences in relationships across populations or for evaluating replication studies.

## Summary Statistics Approach

When only summary statistics are available, the `compare_cor` function can be used:

```{r}
# Comparing correlation r1=0.8 from n=40 with r2=0.2 from n=100
compare_cor(r1 = .8,
            df1 = 38,  # df = n-2
            r2 = .2,
            df2 = 98)  # df = n-2
```

The `compare_cor` function supports different methods for comparing correlations:

```{r}
# Testing equivalence using Fisher's method
compare_cor(r1 = .8,
            df1 = 38,
            r2 = .2,
            df2 = 98,
            null = .2,
            method = "f", # Fisher (can also use "fisher")
            alternative = "e") # Equivalence test
```

Available methods include:

* **Fisher's z transformation** (`method = "fisher"` or `"f"`): Tests the difference between correlations on the z-transformed scale. This is generally recommended for most applications.
* **Kraatz's method** (`method = "kraatz"` or `"k"`): Directly measures the difference between correlation coefficients.

While both methods are appropriate for general significance testing, they may have limited statistical power in some scenarios [@counsell2015equ].

## Bootstrapped Comparison

When raw data is available for both correlations, the `boot_compare_cor` function offers a more robust approach through bootstrapping:

```{r}
set.seed(8922) # Setting seed for reproducibility
# Generating example data
x1 = rnorm(40)
y1 = rnorm(40)

x2 = rnorm(100)
y2 = rnorm(100)

# Bootstrap comparison with winsorized correlation
boot_compare_cor(
  x1 = x1,
  x2 = x2,
  y1 = y1,
  y2 = y2,
  null = .2,
  alternative = "e", # Equivalence test
  method = "win" # Winsorized correlation
)
```


This approach has several advantages:

- It does not rely on the Fisher's z-transformation approximation
- It can incorporate robust correlation methods
- It can provide more accurate confidence intervals, especially when typical assumptions are violated


# Practical Recommendations

When choosing which correlation method to use in TOSTER:

1. **If raw data is available:**
   - For most cases, use `boot_cor_test` with Pearson, Spearman, or Kendall methods
   - When outliers or distribution assumptions are concerns, consider the robust methods (winsorized or percentage bend)

2. **If only summary statistics are available:**
   - Use `corsum_test` for single correlation analysis
   - Use `compare_cor` with the Fisher method for comparing correlations

3. **For equivalence testing:**
   - Carefully select meaningful boundaries (null values) based on your research context
   - Consider what effect size would be practically insignificant in your field


# References
