---
title: "Robust TOST Procedures"
author: "Aaron R. Caldwell"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: True
editor_options: 
  chunk_output_type: console
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Robust TOST Procedures}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

# Introduction to Robust TOST Methods

The Two One-Sided Tests (TOST) procedure is a statistical approach used to test for equivalence between groups or conditions. Unlike traditional null hypothesis significance testing (NHST) which aims to detect differences, TOST is designed to statistically demonstrate similarity or equivalence within predefined bounds.

While the standard `t_TOST` function in TOSTER provides a parametric approach to equivalence testing, it relies on assumptions of normality and homogeneity of variance. In real-world data analysis, these assumptions are often violated, necessitating more robust alternatives. This vignette introduces several robust TOST methods available in the TOSTER package that maintain their validity under a wider range of data conditions.

## When to Use Robust TOST Methods

Consider using the robust alternatives to `t_TOST` when:

- Your data shows notable departures from normality (e.g., skewed distributions, heavy tails)
- Potential outliers are a concern 
- Sample sizes are small or unequal
- You have concerns about violating the assumptions of parametric tests
- You want to confirm parametric test results with complementary nonparametric approaches

The following table provides a quick overview of the robust methods covered in this vignette:

| Method | Function | Key Characteristics | Best Used When |
|--------|----------|---------------------|---------------|
| Wilcoxon TOST | `wilcox_TOST()` | Rank-based, nonparametric | Data is ordinal or non-normal |
| Brunner-Munzel | `brunner_munzel()` with `simple_htest()` | Probability-based, robust to heteroscedasticity | Distribution shapes differ between groups |
| Permutation t-test | `perm_t_test()` | Resampling without replacement, exact p-values possible | Very small samples, exact inference needed |
| Bootstrap TOST | `boot_t_TOST()` | Resampling with replacement, flexible | Moderate samples, visualization desired |
| Log-Transformed TOST | `log_TOST()` | Ratio-based, for multiplicative comparisons | Comparing relative differences (e.g., bioequivalence) |

# Wilcoxon-Mann-Whitney TOST

The Wilcoxon group of tests (includes Mann-Whitney U-test) provide a non-parametric test of differences between groups, or within samples, based on ranks. This provides a test of location shift, which is a fancy way of saying differences in the center of the distribution (i.e., in parametric tests the location is mean). With TOST, there are two separate tests of directional location shift to determine if the location shift is within (equivalence) or outside (minimal effect). The exact calculations can be explored via the documentation of the `wilcox.test` function.

## Key Features and Usage

TOSTER's version is the `wilcox_TOST` function. Overall, this function operates extremely similar to the `t_TOST` function. However, the standardized mean difference (SMD) is *not* calculated. Instead the rank-biserial correlation is calculated for *all* types of comparisons (e.g., two sample, one sample, and paired samples). Also, there is no plotting capability at this time for the output of this function.

The `wilcox_TOST` function is particularly useful when:

- You're working with ordinal data
- You're concerned about outliers influencing your results
- You need a test that makes minimal assumptions about the underlying distributions

### Understanding the Equivalence Bounds (`eqb`)

The equivalence bounds (`eqb`) are specified on the **raw scale** of your data, just like the `mu` argument in `stats::wilcox.test()`. This means `eqb` represents actual differences in location (e.g., location shifts) in the original units of measurement, not standardized effect sizes. Unlike `t_TOST`, where `eqb` *can* be set in terms of standardized mean differences (e.g., Cohen's d), in `wilcox_TOST`, you *must* define `eqb` based on the scale of your data for `wilcox_TOST`.

**The `eqb` and `ses` parameters are completely independent:**

- `eqb` sets the equivalence bounds on the raw data scale for the statistical test
- `ses` only determines which *type* of standardized effect size is calculated and reported (e.g., rank-biserial correlation, concordance probability, or odds)
- Changing `ses` does **not** change the meaning or interpretation of `eqb`

As an example, we can use the sleep data to make a non-parametric comparison of equivalence.

```{r}
data('sleep')
library(TOSTER)

test1 = wilcox_TOST(formula = extra ~ group,
                      data = sleep,
                      paired = FALSE,
                      eqb = .5)


print(test1)
```

### Interpreting the Results

When interpreting the output of `wilcox_TOST`, pay attention to:

1. The p-values for both one-sided tests (`p1` and `p2`)
2. The equivalence test p-value (`TOSTp`), which should be < alpha to claim equivalence
3. The rank-biserial correlation (`rb`) and its confidence interval 

A statistically significant equivalence test (p < alpha) indicates that the observed effect is statistically within your specified equivalence bounds. The rank-biserial correlation provides a measure of effect size, with values ranging from -1 to 1:

- Values near 0 indicate negligible association
- Values around ±0.3 indicate a small effect
- Values around ±0.5 indicate a moderate effect
- Values around ±0.7 or greater indicate a large effect

## Rank-Biserial Correlation

The standardized effect size reported for the `wilcox_TOST` procedure is the rank-biserial correlation. This is a fairly intuitive measure of effect size which has the same interpretation of the common language effect size [@Kerby_2014]. However, instead of assuming normality and equal variances, the rank-biserial correlation calculates the number of favorable (positive) and unfavorable (negative) pairs based on their respective ranks. 

For the two sample case, the correlation is calculated as the proportion of favorable pairs minus the unfavorable pairs.

$$
r_{biserial} = f_{pairs} - u_{pairs}
$$

Where:
- $f_{pairs}$ is the proportion of favorable pairs
- $u_{pairs}$ is the proportion of unfavorable pairs

For the one sample or paired samples cases, the correlation is calculated with ties (values equal to zero) not being dropped. This provides a *conservative* estimate of the rank biserial correlation.

It is calculated in the following steps wherein $z$ represents the values or difference between paired observations:

1. Calculate signed ranks:

$$
r_j = -1 \cdot sign(z_j) \cdot rank(|z_j|)
$$

Where:
- $r_j$ is the signed rank for observation $j$
- $sign(z_j)$ is the sign of observation $z_j$ (+1 or -1)
- $rank(|z_j|)$ is the rank of the absolute value of observation $z_j$

2. Calculate the positive and negative sums:

$$
    R_{+} = \sum_{1\le i \le n, \space z_i > 0}r_j 
$$

$$
    R_{-} = \sum_{1\le i \le n, \space z_i < 0}r_j 
$$

Where:
- $R_{+}$ is the sum of ranks for positive observations
- $R_{-}$ is the sum of ranks for negative observations

3. Determine the smaller of the two rank sums:

$$
T = min(R_{+}, \space R_{-})
$$

$$
S = \begin{cases} -4 & R_{+} \ge R_{-} \\ 4 & R_{+} < R_{-} \end{cases}
$$

Where:
- $T$ is the smaller of the two rank sums
- $S$ is a sign factor based on which rank sum is smaller

4. Calculate rank-biserial correlation:

$$
r_{biserial} = S \cdot | \frac{\frac{T - \frac{(R_{+} + R_{-})}{2}}{n}}{n + 1} |
$$

Where:
- $n$ is the number of observations (or pairs)
- The final value ranges from -1 to 1

## Confidence Intervals

The Fisher approximation is used to calculate the confidence intervals.

For paired samples, or one sample, the standard error is calculated as the following:

$$
SE_r = \sqrt{ \frac {(2 \cdot nd^3 + 3 \cdot nd^2 + nd) / 6} {(nd^2 + nd) / 2} }
$$

wherein, nd represents the total number of observations (or pairs).

For independent samples, the standard error is calculated as the following:

$$
SE_r = \sqrt{\frac {(n1 + n2 + 1)} { (3 \cdot n1 \cdot n2)}}
$$

Where:

- $n1$ and $n2$ are the sample sizes of the two groups

The confidence intervals can then be calculated by transforming the estimate.

$$
r_z = atanh(r_{biserial})
$$

Then the confidence interval can be calculated and back transformed.

$$
r_{CI} = tanh(r_z  \pm  Z_{(1 - \alpha / 2)} \cdot SE_r)
$$

Where:

- $Z_{(1 - \alpha / 2)}$ is the critical value from the standard normal distribution
- $\alpha$ is the significance level (typically 0.05)

## Conversion to other effect sizes

Two other effect sizes can be calculated for non-parametric tests. First, there is the concordance probability, which is also known as the c-statistic, c-index, or probability of superiority^[Directly inspired by this blog post from Professor Frank Harrell https://hbiostat.org/blog/post/wpo/]. The c-statistic is converted from the correlation using the following formula:

$$
c = \frac{(r_{biserial} + 1)}{2}
$$

The c-statistic can be interpreted as the probability that a randomly selected observation from one group will be greater than a randomly selected observation from another group. A value of 0.5 indicates no difference between groups, while values approaching 1 indicate perfect separation between groups.

The Wilcoxon-Mann-Whitney odds [@wmwodds], also known as the "Generalized Odds Ratio" [@agresti], is calculated by converting the c-statistic using the following formula:

$$
WMW_{odds} = e^{logit(c)}
$$

Where $logit(c) = \ln\frac{c}{1-c}$

The WMW odds can be interpreted similarly to a traditional odds ratio, representing the odds that an observation from one group is greater than an observation from another group.

Either effect size is available by simply modifying the `ses` argument for the `wilcox_TOST` function. **Note:** In all examples below, `eqb = .5` remains on the raw data scale (the scale of the `extra` variable in hours of sleep). The `ses` argument only changes which standardized effect size is calculated and reported; it does not change the equivalence bounds or the hypothesis test itself.

```{r}
# Rank biserial
wilcox_TOST(formula = extra ~ group,
                      data = sleep,
                      paired = FALSE,
                      ses = "r",
                      eqb = .5)

# Odds

wilcox_TOST(formula = extra ~ group,
                      data = sleep,
                      paired = FALSE,
                      ses = "o",
                      eqb = .5)

# Concordance

wilcox_TOST(formula = extra ~ group,
                      data = sleep,
                      paired = FALSE,
                      ses = "c",
                      eqb = .5)

```

### Guidelines for Selecting Effect Size Measures

- **Rank-biserial correlation (`"r"`)** is useful when you want a correlation-like measure that's easily interpretable and comparable to other correlation coefficients.
- **Concordance probability (`"c"`)** is beneficial when you want to express the effect in terms of probability, making it accessible to non-statisticians.
- **WMW odds (`"o"`)** is helpful when you want to express the effect in terms familiar to those who work with odds ratios in logistic regression or epidemiology.

# Brunner-Munzel

As @karch2021 explained, there are some reasons to dislike the WMW family of tests as the non-parametric alternative to the t-test. Regardless of the underlying statistical arguments^[I would like to note that I think the statistical properties of the WMW tests are sound, and Frank Harrell has written [many](https://www.fharrell.com/post/po/) [blogposts](https://www.fharrell.com/post/wpo/) outlined their sound application in biomedicine. ], it can be argued that the interpretation of the WMW tests, especially when involving equivalence testing, is a tad difficult.
Some may want a non-parametric alternative to the WMW test, and the Brunner-Munzel test(s) may be a useful option.

## Overview and Advantages

The Brunner-Munzel test [@brunner2000; @neubert2007] offers several advantages over the Wilcoxon-Mann-Whitney tests:

1. It does not assume equal distributions (shapes) between groups
2. It is robust to heteroscedasticity (unequal variances)
3. It provides a more interpretable effect size measure ("relative effect")
4. It maintains good statistical properties even with unequal sample sizes

The Brunner-Munzel test is based on calculating the "stochastic superiority" [@karch2021, i.e., probability of superiority], which is usually referred to as the relative effect, based on the ranks of the two groups being compared (X and Y). A Brunner-Munzel type test is then a directional test of an effect, and answers a question akin to "what is the probability that a randomly sampled value of X will be greater than Y?"

$$
\hat p = P(X>Y) + 0.5 \cdot P(X=Y)
$$

Where:

- $P(X>Y)$ is the probability that a random value from X exceeds a random value from Y
- $P(X=Y)$ is the probability that random values from X and Y are equal
- The 0.5 factor means ties contribute half their weight to the probability

The relative effect $\hat p$ has an intuitive interpretation:

- $\hat p = 0.5$ indicates no difference between groups
- $\hat p > 0.5$ indicates values in X tend to be greater than values in Y
- $\hat p < 0.5$ indicates values in X tend to be smaller than values in Y

## Basics of the Calculative Approach

In this section, I will quickly detail the calculative approach that underlies the Brunner-Munzel test in `TOSTER`.

A studentized test statistic can be calculated as:

$$
t = \sqrt{N} \cdot \frac{\hat p -p_{null}}{s}
$$

Where:

- $N$ is the total sample size
- $\hat p$ is the estimated relative effect
- $p_{null}$ is the null hypothesis value (typically 0.5)
- $s$ is the rank-based Brunner-Munzel standard error

The default null hypothesis $p_{null}$ is typically 0.5 (50% probability of superiority is the default null), and $s$ refers the rank-based Brunner-Munzel standard error. The null can be modified therefore allowing for equivalence testing *directly based on the relative effect*. Additionally, for paired samples the probability of superiority is based on a *hypothesis of exchangability* and is not based on the differences scores^[This means the relative effect will *not* match the concordance probability provided by `ses_calc`].

For more details on the calculative approach, I suggest reading @karch2021. At small sample sizes, it is recommended that the permutation version of the test (`perm = TRUE`) be used rather than the basic test statistic approach.

## Example

The interface for the function is very similar to the `t.test` function. The `brunner_munzel` function itself does not allow for equivalence tests, but you can set an alternative hypothesis for "two.sided", "less", or "greater".

```{r}
# studentized test
brunner_munzel(formula = extra ~ group,
                      data = sleep,
                      paired = FALSE)
# permutation
brunner_munzel(formula = extra ~ group,
                      data = sleep,
                      paired = FALSE,
               perm = TRUE)
```

The `simple_htest` function allows TOST tests using a Brunner-Munzel test by setting the alternative to "equivalence" or "minimal.effect". The equivalence bounds, based on the relative effect, can be set with the `mu` argument.

```{r}
# permutation based Brunner-Munzel test of equivalence
simple_htest(formula = extra ~ group,
             test = "brunner",
             data = sleep,
             paired = FALSE,
             alternative = "equ",
             mu = .7,
             perm = TRUE)

```

### Interpreting Brunner-Munzel Results

When interpreting the Brunner-Munzel test results:

1. The relative effect (p-hat) is the primary measure of interest
2. For equivalence testing, you're testing whether this relative effect falls within your specified bounds
3. A significant equivalence test suggests that the probability of superiority is statistically confined to your specified range

### When to Use Permutation

The permutation approach (`perm = TRUE`) is recommended when:

- Sample sizes are small (generally n < 30 per group)
- Data distributions are highly skewed or contain outliers
- Precision is more important than computational speed

Note that the permutation approach can be computationally intensive for large datasets, potentially increasing processing time significantly. Additionally, with a permutation test you may observe situations where the confidence interval and the p-values would yield *different* conclusions.

# Resampling Methods: Bootstrapping and Permutation

Resampling methods provide robust alternatives to parametric t-tests by using the data itself to approximate the sampling distribution of the test statistic. TOSTER offers two complementary resampling approaches: bootstrapping and permutation testing. While both methods relax distributional assumptions, they differ in their mechanics and are suited to different situations.

## Conceptual Overview

**Bootstrapping** resamples *with replacement* from the observed data to estimate the variability of a statistic. The bootstrap distribution approximates what we would see if we could repeatedly sample from the true population. This approach is particularly useful when you want confidence intervals and effect size estimates alongside your p-values.

**Permutation testing** resamples *without replacement* by randomly reassigning observations to groups (or flipping signs for one-sample/paired tests). Under the null hypothesis of no difference, group labels are exchangeable, and the permutation distribution represents the exact null distribution. Permutation tests provide valid inference even with very small samples where asymptotic approximations may fail.

## When to Use Each Method

| Scenario | Recommended Method | Rationale |
|----------|-------------------|-----------|
| Very small samples (paired/one-sample N < 17, two-sample N < 20) | Permutation | Exact permutations are computationally feasible and provide exact p-values |
| Moderate to large samples | Bootstrap | More computationally efficient; provides stable CI estimates |
| Outliers present | Permutation with trimming | Combines exact inference with robust location estimation |
| Complex data structures | Bootstrap | More flexible for non-standard designs |
| Publication requiring exact p-values | Permutation (if feasible) | No Monte Carlo error when exact permutations computed |

The sample size thresholds above reflect computational practicality. For one-sample or paired tests, the number of possible sign permutations is $2^n$, which equals 65,536 when n = 16 and jumps to 131,072 at n = 17. For two-sample tests, the number of possible group assignments is $\binom{n_x + n_y}{n_x}$, which grows even faster. When exact enumeration is feasible, permutation tests yield exact p-values with no Monte Carlo error.

# Permutation t-test

The `perm_t_test` function implements studentized permutation tests following the approaches of @janssen1997 and @chung2013. The studentized approach computes a t-statistic for each permutation, making the test valid even under heteroscedasticity (unequal variances).

## Permutation Procedure

For **one-sample and paired tests**, permutation is performed by randomly flipping the signs of the centered observations (or difference scores). Under the null hypothesis that the mean equals the hypothesized value, positive and negative deviations are equally likely, making sign assignments exchangeable.

For **two-sample tests**, permutation is performed by randomly reassigning observations to the two groups. Under the null hypothesis of no group difference, group membership is arbitrary, making the labels exchangeable.

## Combining Permutation with Trimmed Means

A key feature of `perm_t_test` is its support for trimmed means via the `trim` argument. When `trim > 0`, the function uses Yuen's approach: trimmed means for location estimation and winsorized variances for standard error calculation. This combination is particularly powerful because it provides robustness against outliers (through trimming) together with exact or near-exact inference (through permutation).

Trimming is helpful when:

- Data contain outliers or extreme values that would unduly influence the mean
- Distributions have heavy tails
- You want robust location estimates without sacrificing the exactness of permutation inference

A common choice is `trim = 0.1` (10% trimming) or `trim = 0.2` (20% trimming), though the optimal amount depends on the suspected degree of contamination.

## Example: Basic Permutation Test

```{r}
data('sleep')

# Two-sample permutation t-test
perm_result <- perm_t_test(extra ~ group, 
                           data = sleep,
                           R = 1999)
perm_result
```

## Example: Permutation Test with Trimming

When outliers are a concern, combining permutation with trimming provides doubly robust inference:

```{r}
# Simulate data with outliers
set.seed(42)
x <- c(rnorm(18, mean = 0), 8, 12)  # Two outliers
y <- rnorm(20, mean = 0)

# Standard permutation test (sensitive to outliers)
perm_t_test(x, y, R = 999)

# Trimmed permutation test (robust to outliers)
perm_t_test(x, y, trim = 0.1, R = 999)
```

## Example: Equivalence Testing with Permutation

The `perm_t_test` function supports TOSTER's equivalence and minimal effect alternatives:

```{r}
# Equivalence test: is the effect within ±3 units?
perm_t_test(extra ~ group, 
            data = sleep,
            alternative = "equivalence",
            mu = c(-3, 3),
            R = 999)
```

## Exact vs. Monte Carlo Permutations

When the total number of possible permutations is less than or equal to `R`, the function automatically computes all exact permutations and prints a message. For example, with a paired sample of n = 10, there are $2^{10} = 1024$ possible sign permutations, so requesting `R = 1999` would trigger exact computation.

```{r}
# Small paired sample - exact permutations will be computed
before <- c(5.1, 4.8, 6.2, 5.7, 6.0, 5.5, 4.9, 5.8)
after <- c(5.6, 5.2, 6.7, 6.1, 6.5, 5.8, 5.3, 6.2)

perm_t_test(x = before, y = after,
            paired = TRUE,
            alternative = "less",
            R = 999)
```

# Bootstrap t-test

The `boot_t_TOST` function provides bootstrap-based inference using the percentile bootstrap approach outlined by @efron93 (see chapter 16). The bootstrapped p-values are derived from the studentized version of a test of mean differences. Overall, the results should be similar to `t_TOST` but with greater robustness to distributional violations.

## Advantages of Bootstrapping

Bootstrap methods offer several advantages for equivalence testing:

1. They make minimal assumptions about the underlying data distribution
2. They are robust to deviations from normality
3. They provide realistic confidence intervals even with small samples
4. They can handle complex data structures and dependencies
5. They often provide more accurate results when parametric assumptions are violated

## Bootstrap Algorithm (Two-Sample Case)

1. Form B bootstrap data sets from x* and y* wherein x* is sampled with replacement from $\tilde x_1,\tilde x_2, ... \tilde x_n$ and y* is sampled with replacement from $\tilde y_1,\tilde y_2, ... \tilde y_n$

   Where:
   - B is the number of bootstrap replications (set using the `R` parameter)
   - $\tilde x_i$ and $\tilde y_i$ represent the original observations in each group

2. t is then evaluated on each sample, but the mean of each sample (y or x) and the overall average (z) are subtracted from each 

$$
t(z^{*b}) = \frac {(\bar x^*-\bar x - \bar z) - (\bar y^*-\bar y - \bar z)}{\sqrt {sd_y^*/n_y + sd_x^*/n_x}}
$$

Where:

- $\bar x^*$ and $\bar y^*$ are the means of the bootstrap samples
- $\bar x$ and $\bar y$ are the means of the original samples
- $\bar z$ is the overall mean
- $sd_x^*$ and $sd_y^*$ are the standard deviations of the bootstrap samples
- $n_x$ and $n_y$ are the sample sizes

3. An approximate p-value can then be calculated as the number of bootstrapped results greater than the observed t-statistic from the sample.

$$
p_{boot} = \frac {\#t(z^{*b}) \ge t_{sample}}{B}
$$

Where:
- $\#t(z^{*b}) \ge t_{sample}$ is the count of bootstrap t-statistics that exceed the observed t-statistic
- B is the total number of bootstrap replications

The same process is completed for the one sample case but with the one sample solution for the equation outlined by $t(z^{*b})$. The paired sample case in this bootstrap procedure is equivalent to the one sample solution because the test is based on the difference scores.

## Choosing the Number of Bootstrap Replications

When using bootstrap methods, the choice of replications (the `R` parameter) is important:

- **R = 999 or 1999**: Recommended for standard analyses
- **R = 4999 or 9999**: Recommended for publication-quality results or when precise p-values are needed
- **R < 500**: Acceptable only for exploratory analyses or when computational resources are limited

Larger values of R provide more stable results but increase computation time. For most purposes, 999 or 1999 replications strike a good balance between precision and computational efficiency.

## Example

We can use the sleep data to see the bootstrapped results. Notice that the plots show how the re-sampling via bootstrapping indicates the instability of Hedges's d~z~.

```{r}
data('sleep')

test1 = boot_t_TOST(formula = extra ~ group,
                      data = sleep,
                      paired = TRUE,
                      eqb = .5,
                    R = 499)


print(test1)

plot(test1)
```

### Interpreting Bootstrap TOST Results

When interpreting the results of `boot_t_TOST`:

1. The bootstrap p-values (`p1` and `p2`) represent the empirical probability of observing the test statistic or more extreme values under repeated sampling
2. The confidence intervals are derived directly from the empirical distribution of bootstrap samples
3. The distribution plots provide visual insight into the variability of the effect size estimate

For equivalence testing, examine whether both bootstrap p-values are significant (< alpha) and whether the confidence interval for the effect size falls entirely within the equivalence bounds.

# Comparing Bootstrap and Permutation Approaches

To illustrate the similarities and differences between these methods, let's apply both to the same dataset:

```{r}
# Same equivalence test using both methods
data('sleep')

# Bootstrap approach
boot_result <- boot_t_test(extra ~ group, 
                           data = sleep,
                           alternative = "equivalence",
                           mu = c(-2, 2),
                           R = 999)

# Permutation approach  
perm_result <- perm_t_test(extra ~ group,
                           data = sleep,
                           alternative = "equivalence", 
                           mu = c(-2, 2),
                           R = 999)

# Compare results
boot_result
perm_result
```

Both methods test the same hypothesis and should yield similar conclusions. Key differences to note:

- **P-values**: May differ slightly due to Monte Carlo variability and the different resampling mechanisms
- **Confidence intervals**: Bootstrap uses the resampled effect distribution; permutation uses the percentile method on permuted effects
- **Exact inference**: When sample sizes are small enough, permutation provides exact p-values with no Monte Carlo error

## Summary: Choosing Between Methods

| Feature | Permutation (`perm_t_test`) | Bootstrap (`boot_t_TOST`) |
|---------|----------------------------|---------------------------|
| Resampling type | Without replacement | With replacement |
| Exact p-values possible | Yes (small samples) | No |
| Trimmed means support | Yes (`trim` argument) | No |
| Effect size plots | No | Yes |
| Best for | Small samples, exact inference | Moderate samples, visualization |

For very small samples where exact permutations are feasible, permutation testing is generally preferred because it provides exact p-values (i.e., no need to set a seed). For larger samples or when you want the visualization capabilities of `boot_t_TOST`, bootstrapping is a good choice. When outliers are a concern, using `perm_t_test` with trimming combines the benefits of exact inference with robust location estimation.

# Ratio of Difference (Log Transformed)

In many bioequivalence studies, the differences between drugs are compared on the log scale [@he2022].
The log scale allows researchers to compare the ratio of two means.

$$
log ( \frac{y}{x} ) = log(y) - log(x)
$$

Where:
- y and x are the means of the two groups being compared
- The transformation converts multiplicative relationships into additive ones

## Why Use The Natural Log Transformation?

The [United States Food and Drug Administration (FDA)](https://www.fda.gov/regulatory-information/search-fda-guidance-documents/bioavailability-and-bioequivalence-studies-submitted-ndas-or-inds-general-considerations)^[Food and Drug Administration (2014). Bioavailability and Bioequivalence Studies Submitted in NDAs or INDs — General Considerations.Center for Drug Evaluation and Research. Docket: FDA-2014-D-0204] has stated a rationale for using the log transformed values:

> Using logarithmic transformation, the general linear statistical model employed in the analysis of BE data allows inferences about the difference between the two means on the log scale, which can then be retransformed into inferences about the ratio of the two averages (means or medians) on the original scale. Logarithmic transformation thus achieves a general comparison based on the ratio rather than the differences. 

Log transformation offers several advantages:

1. It facilitates the analysis of **relative** rather than absolute differences
2. It often makes right-skewed distributions more symmetric
3. It stabilizes variance when variability increases with the mean
4. It provides an easy-to-interpret interpretable effect size (ratio of means)

In addition, the FDA considers two drugs as bioequivalent when the ratio between x and y is less than 1.25 and greater than 0.8 (1/1.25), which is the default equivalence bound for the log functions.

## Applications Beyond Bioequivalence

While log transformation is standard in bioequivalence studies, it's useful in many other contexts:

- **Economics**: Comparing percentage changes in economic indicators
- **Environmental science**: Analyzing concentration ratios of pollutants
- **Biology**: Examining growth rates or concentration ratios
- **Medicine**: Comparing relative efficacy of treatments
- **Psychology**: Analyzing response time ratios

Consider using log transformation whenever your research question is about relative rather than absolute differences, particularly when the data follow a multiplicative rather than additive pattern.

## log_TOST

For example, we could compare whether the cars of different transmissions
are "equivalent" with regards to gas mileage. 
We can use the default equivalence bounds (`eqb = 1.25`).

```{r}
log_TOST(
  mpg ~ am,
  data = mtcars
)
```

Note, that the function produces t-tests similar to the `t_TOST` function, but
provides two effect sizes. The means ratio on the log scale (the scale of the test statistics),
and the means ratio. The means ratio is missing standard error because the
confidence intervals and estimate are simply the log scale results exponentiated.

### Interpreting the Means Ratio

When interpreting the means ratio:

- A ratio of 1.0 indicates perfect equivalence (no difference)
- Ratios > 1.0 indicate that the first group has higher values than the second
- Ratios < 1.0 indicate that the first group has lower values than the second

For equivalence testing with the default bounds (0.8, 1.25):
- Equivalence is established when the 90% confidence interval for the ratio falls entirely within (0.8, 1.25)
- This range corresponds to a difference of ±20% on a relative scale

## Bootstrap + Log

However, it has been noted in the statistics literature that t-tests on the
logarithmic scale can be biased, and it is recommended that bootstrapped tests
be utilized instead. Therefore, the `boot_log_TOST` function can be utilized to
perform a more precise test.

```{r}
boot_log_TOST(
  mpg ~ am,
  data = mtcars,
  R = 499
)
```

The bootstrapped version is particularly recommended when:

- Sample sizes are small (n < 30 per group)
- Data show notable deviations from log-normality
- You want to ensure robust confidence intervals

# Just Estimate an Effect Size

It was requested that a function be provided that only calculates a robust effect size.
Therefore, I created the `ses_calc` and `boot_ses_calc` functions as robust effect size calculation^[The results differ greatly because the bootstrap CI method, basic bootstrap, is more conservative than the parametric method. This difference is more apparent with extremely small samples like that in the `sleep` dataset.].
The interface is almost the same as `wilcox_TOST` but you don't set an equivalence bound.

```{r}
ses_calc(formula = extra ~ group,
         data = sleep,
         paired = TRUE,
         ses = "r")

# Setting bootstrap replications low to
## reduce compiling time of vignette
boot_ses_calc(formula = extra ~ group,
         data = sleep,
         paired = TRUE,
         R = 199,
         boot_ci = "perc", # recommend percentile bootstrap for paired SES
         ses = "r") 
```

## Choosing Between Different Bootstrap CI Methods

The `boot_ses_calc` function offers several bootstrap confidence interval methods through the `boot_ci` parameter:

- **"perc"** (Percentile): Simple and intuitive, works well for symmetric distributions
- **"basic"**: Similar to percentile but adjusts for bias, more conservative
- **"norm"** (Normal approximation): Assumes normality of the bootstrap distribution


# Summary Comparison of Robust TOST Methods

| Method | Key Advantages | Limitations | Best Use Cases |
|--------|---------------|-------------|---------------|
| **Wilcoxon TOST** | Simple, widely accepted, minimal assumptions | Less power than parametric tests with normal data | Ordinal data, non-normal distributions, presence of outliers |
| **Brunner-Munzel** | Robust to unequal distributions, interpretable effect | Computationally intensive with permutation | Different distribution shapes between groups, heteroscedasticity |
| **Permutation t-test** | Exact p-values (small samples), supports trimmed means | Computationally intensive for large samples | Very small samples (N < 17 paired, N < 20 two-sample), outliers with trimming |
| **Bootstrap TOST** | Flexible, visualization of effect distributions | Results vary slightly between runs, no trimming option | Moderate samples, when effect size plots are desired |
| **Log-Transformed** | Focuses on relative differences, often stabilizes variance | Requires positive data, can be biased with small samples | Bioequivalence studies, comparing ratios rather than absolute differences |

# Conclusion

The robust TOST procedures provided in the TOSTER package offer reliable alternatives to standard parametric equivalence testing when data violate typical assumptions. By selecting the appropriate robust method for your specific data characteristics and research question, you can ensure more valid statistical inferences about equivalence or minimal effects.

Remember that no single method is universally superior - the choice depends on your data structure, sample size, and specific research question. When in doubt, running multiple approaches and comparing results can provide valuable insights into the robustness of your conclusions.

# References
