---
title: "TOSTER validation against published literature"
author: "Daniel Lakens"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TOSTER validation against published literature}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

#**Tests against Literature, existing packages, and MiniTab**

Below I reproduce some examples from the literature, compare TOSTER against other R packages that have equivalence test functions, and against Minitab. MiniTab software has implemented equivalence tests and power analysis for one-sample, independent, and paired t-tests. It is excellent - by far the most user-friendly software for anyone who prefers a commercial statistics program over an open source solution (it also has equivalence tests for 2x2 crossover designs, but not for correlations or meta-analysis). It also allows users to enter summary statistics, which is used here to compare the output of the equivalence functions in TOSTER against minitab. Note that when SD  = 1, the mean difference equals the standardized difference. The MiniTab output is followed by the R function.


#**TOST for two independent samples**


##*Example 1.1*

From page 196 of Stegner, B. L., Bostrom, A. G., & Greenfield, T. K. (1996). Equivalence testing for use in psychosocial and services research: An introduction with examples. Evaluation and Program Planning, 19(3), 193-198. https://doi.org/10.1016/0149-7189(96)00011-0

Randomized controlled trial of two conditions of case management, with the main dependent variable being ratings on the Brief Psychiatric Rating Scale (BPRS). The means, sd's, and sample sizes are: 

Control: M = 1.5679, SD = 0.4285, n = 64
Experimental: M = 1.6764, SD = 0.4748, n = 70

Calculated t-values for TOST are t1 = 5.383 and t2 = -2.616

These results can be reproduced with TOSTER package and the TOSTtwo.raw R code: 

```{r}
require(TOSTER)
TOSTtwo.raw(m1=1.5679,m2=1.6764,sd1=0.4285,sd2=0.4748,n1=64,n2=70,low_eqbound=-0.3136, high_eqbound=0.3136,alpha=0.05, var.equal=TRUE)

```


##*Example 1.2*

From page 225 of Limentani, G. B., Ringo, M. C., Ye, F., Bergquist, M. L., & MCSorley, E. O. (2005). Beyond the t-test: statistical equivalence testing. Analytical Chemistry, 77(11), 221-A.

Reproducing the 90% CI and significant equivalence test.

```{r}
TOSTtwo.raw(m1=89.3,m2=87.7,sd1=1.9,sd2=1.3,n1=12,n2=12,low_eqbound=-3.7, high_eqbound=3.7,alpha=0.05, var.equal=TRUE)

```


#*Example 1.3*

From Meyners, M. (2012). Equivalence tests - A review. Food Quality and Preference, 26(2), 231-245. https://doi.org/10.1016/j.foodqual.2012.05.003, page 237:

To illustrate the approach, we consider the data presented by Bi (2005) of 100 consumers each from two cities A and B, giving overall liking scores on a 9-point scale. The equivalence margin is set to d = 0.5, i.e. a difference of 0.5 or less in mean values is considered negligible. With mean values of lA = 7.1 and lB = 6.9 and variances s2A = 2.0 and s2B = 2.2, the corresponding t-statistic is 0.976, and the
p-values for the two one-sided test problems become p1 = 0.0717 and p2 = 0.0003.

We can use the TOST procedure for an independent t-test to reproduce the p-values and 90% CI:

```{r}
TOSTtwo.raw(m1=7.1, m2=6.9, sd1=sqrt(2.0), sd2=sqrt(2.2), n1=100, n2=100,low_eqbound=-0.5, high_eqbound=0.5,alpha=0.05, var.equal=TRUE)

```
This outcome matches the result of (page 237): The 90% CI is [0.139; 0.539] with a p-value of approximately 0.07.  


##*Example 1.4*

From Table 1 in Quertemont, E. (2011). How to Statistically Show the Absence of an Effect. Psychologica Belgica, 51(2), 109. https://doi.org/10.5334/pb-51-2-109

Example 1: t(8) = 1.52; p = 0.92
Example 2: t(298) = -2.23; p = 0.013

```{r}
#Example 1
TOSTtwo.raw(m1=459.09, m2=402.61, sd1=47.53, sd2=38.42, n1=5, n2=5, low_eqbound=-15, high_eqbound=15,alpha=0.05, var.equal=TRUE)

#Example 2
TOSTtwo.raw(m1=407.24, m2=402.98, sd1=41.77, sd2=41.79, n1=150, n2=150, low_eqbound=-15, high_eqbound=15,alpha=0.05, var.equal=TRUE)

```

These results perfectly match those in Table 1 of Quertemont (2011).


##*Example 1.5*

MiniTab by default uses Welch's t-test when testing for equivalence, which does not require the assumption of equal variances, which matters with unequal sample sizes. TOSTER also presents Welch's t-test by default. Note that minitab rounds degrees of freedom down to a whole number. This is a conservative procedure. TOSTER does not. The difference is rarely noticable. 

### MiniTab Result Two-Sample Equivalence Test 

*Method

Test mean = mean of Test
Reference mean = mean of Reference
Equal variances were not assumed for the analysis.


Descriptive Statistics

Variable     N  Mean  StDev  SE Mean
Test        86   2,5    1,3  0,14018
Reference  102   2,3    1,1  0,10892


Difference: Mean(Test) - Mean(Reference)

Difference       SE         95% CI         Equivalence Interval
   0,20000  0,17752  (-0,093626; 0,49363)       (-0,4; 0,4)

CI is not within the equivalence interval. Cannot claim equivalence.


Test

Null hypothesis:         Difference < or = -0,4 or Difference > or = 0,4
Alternative hypothesis:  -0,4 < Difference < 0,4
alpha level:                 0,05

Null Hypothesis     DF  T-Value  P-Value
Difference < or = -0,4  167   3,3799    0,000
Difference > or = 0,4   167  -1,1266    0,131

The greater of the two P-Values is 0,131. Cannot claim equivalence.

###Reproduce results in TOSTER

```{r}
TOSTtwo.raw(m1=2.5,m2=2.3,sd1=1.3, sd2=1.1,n1=86,n2=102,low_eqbound=-0.4, high_eqbound=0.4,alpha=0.05)
```


##*Example 1.6*

Now for the same data as example 1.5, but assuming equal variances:

###Results MiniTab Two-Sample Equivalence Test 

Method

Test mean = mean of Test
Reference mean = mean of Reference
Equal variances were assumed for the analysis.


Descriptive Statistics

Variable     N  Mean  StDev  SE Mean
Test        86   2,5    1,3  0,14018
Reference  102   2,3    1,1  0,10892

Pooled StDev = 1,19556


Difference: Mean(Test) - Mean(Reference)

Difference       SE         95% CI         Equivalence Interval
   0,20000  0,17502  (-0,089331; 0,48933)       (-0,4; 0,4)

CI is not within the equivalence interval. Cannot claim equivalence.


Test

Null hypothesis:         Difference < or = -0,4 or Difference > or = 0,4
Alternative hypothesis:  -0,4 < Difference < 0,4
alpha level:                 0,05

Null Hypothesis     DF  T-Value  P-Value
Difference < or = -0,4  186   3,4281    0,000
Difference > or = 0,4   186  -1,1427    0,127

The greater of the two P-Values is 0,127. Cannot claim equivalence.

###Reproducing results with TOSTER

```{r}
TOSTtwo.raw(m1=2.5,m2=2.3,sd1=1.3, sd2=1.1,n1=86,n2=102,low_eqbound=-0.4, high_eqbound=0.4,alpha=0.05, var.equal=TRUE)
```


```{r}
TOSTtwo.raw(m1=2.5,m2=2.3,sd1=1.3, sd2=1.1,n1=86,n2=102,low_eqbound=-0.4, high_eqbound=0.4,alpha=0.05)
```


#**Paired t-test**

##*Example 2.1*

We can compare the TOSTpaired.raw function against the equivalence package. 

Paired t-test in equivalence package (from example in tost function) and TOSTpaired.raw.

```{r}
require(equivalence)
data(ufc)
#Remove missing data
ufc<-ufc[complete.cases(ufc[,9:10]),]
      
tost(ufc$Height.m.p, ufc$Height.m, epsilon = 1, paired = TRUE)

TOSTpaired.raw(n=length(ufc$Height.m.p),m1=mean(ufc$Height.m.p,na.rm = TRUE), m2=mean(ufc$Height.m,na.rm = TRUE), sd1=sd(ufc$Height.m.p,na.rm = TRUE), sd2=sd(ufc$Height.m,na.rm = TRUE), r12=cor(ufc$Height.m,ufc$Height.m.p, use="pairwise.complete.obs"), low_eqbound=-1,high_eqbound=1, alpha=0.05)
```


##*Example 2.2*

Quertemont also provides an example for paired t-tests in the appendix (page 124):
Mdif = 1, SDif = 1.58, d = .5, t = 0.71, df = 4, p = .74 with a 95% CI from -.96 to + 2.96.

```{r}
morning<-c(3,4,4,5,4)
evening<-c(1,4,1,4,5)

TOSTpaired.raw(n=5,m1=mean(morning), m2=mean(evening), sd1=sd(morning), sd2=sd(evening), r12=cor(morning,evening), low_eqbound=-0.5,high_eqbound=0.5, alpha=0.05)
```

This output matches the calculations in Quertemont (2011).


##*Example 2.3*


MiniTab does not allow users to perform equivalence tests for paired samples t-test based on summary data, but this equivalence test can be calculated based on the raw data. 

Data: 
DV1<-c(4,2,4,3,5,4,3,4,5,4,2,3,4,5,5)
DV2<-c(3,4,2,3,5,4,3,4,5,3,4,3,2,3,4)

###Results MiniTab Equivalence Test with Paired Data: DV1, DV2 

Method

Test mean = mean of DV1
Reference mean = mean of DV2


Descriptive Statistics

Variable   N    Mean    StDev  SE Mean
DV1       15  3,8000   1,0142  0,26186
DV2       15  3,4667  0,91548  0,23637


Difference: Mean(DV1) - Mean(DV2)

Difference    StDev       SE         95% CI        Equivalence Interval
   0,33333  1,23443  0,31873  (-0,22804; 0,89471)       (-0,8; 0,8)

CI is not within the equivalence interval. Cannot claim equivalence.


Test

Null hypothesis:         Difference < or = -0,8 or Difference > or = 0,8
Alternative hypothesis:  -0,8 < Difference < 0,8
alpha level:                 0,05


Null Hypothesis    DF  T-Value  P-Value
Difference < or = -0,8  14   3,5558    0,002
Difference > or = 0,8   14  -1,4642    0,083

The greater of the two P-Values is 0,083. Cannot claim equivalence.

###Reproduce Results with TOSTER

```{r}
DV1<-c(4,2,4,3,5,4,3,4,5,4,2,3,4,5,5)
DV2<-c(3,4,2,3,5,4,3,4,5,3,4,3,2,3,4)

TOSTpaired.raw(n=length(DV1),m1=mean(DV1), m2=mean(DV2), sd1=sd(DV1), sd2=sd(DV2), r12=cor(DV1,DV2), low_eqbound=-0.8,high_eqbound=0.8, alpha=0.05)

```



#**One-sample t-test**


##*Example 3.1*

###Result MiniTab One-Sample Equivalence Test 

Method

Target = 0


Descriptive Statistics

  N  Mean  StDev  SE Mean
100  -0,3      1  0,10000


Difference: Mean - Target

Difference       SE      95% CI     Equivalence Interval
  -0,30000  0,10000  (-0,46604; 0)       (-0,5; 0,5)

CI is within the equivalence interval. Can claim equivalence.


Test

Null hypothesis:         Difference < or = -0,5 or Difference > or = 0,5
Alternative hypothesis:  -0,5 < Difference < 0,5
alpha level:                 0,05

Null Hypothesis    DF  T-Value  P-Value
Difference < or = -0,5  99   2,0000    0,024
Difference > or = 0,5   99  -8,0000    0,000

The greater of the two P-Values is 0,024. Can claim equivalence.

###Reproduce with TOSTER:

```{r}
TOSTone.raw(m=-0.3,mu=0.0,sd=1,n=100,low_eqbound=-0.5, high_eqbound=0.5, alpha=0.05)
```
Which should be identical to the TOSTone test (because when SD = 1, unstandardized scores equal standardized scores)
```{r}
TOSTone(m=-0.3,mu=0.0,sd=1,n=100,low_eqbound_d=-0.5, high_eqbound_d=0.5, alpha=0.05)
```

##*Example 3.2*

###MiniTab Result One-Sample Equivalence Test 

Method

Target = 0


Descriptive Statistics

  N  Mean  StDev  SE Mean
100     1    1,2  0,12000


Difference: Mean - Target

Difference       SE     95% CI    Equivalence Interval
    1,0000  0,12000  (0; 1,1992)       (-0,8; 0,8)

CI is not within the equivalence interval. Cannot claim equivalence.


Test

Null hypothesis:         Difference < or = -0,8 or Difference > or = 0,8
Alternative hypothesis:  -0,8 < Difference < 0,8
alpha level:                 0,05

Null Hypothesis    DF  T-Value  P-Value
Difference < or = -0,8  99   15,000    0,000
Difference > or = 0,8   99   1,6667    0,951

The greater of the two P-Values is 0,951. Cannot claim equivalence.

###Reproducing the results with TOSTER

```{r}
TOSTone.raw(m=-1,mu=0.0,sd=1.2,n=100,low_eqbound=-0.8, high_eqbound=0.8, alpha=0.05)
```



#**Equivalence test for correlations**

##*Example 4.1*

Quertemont (2011) provides an example in the appendix (page 126): Suppose a sample with N = 50, r = .1, and d = .2, and calculated that p(Z < -0.70) = .24. The confidence interval for r is -0.18 to 0.37

We can reproduce this result using the TOSTr function: 

```{r}
TOSTr(n=50, r = 0.1, low_eqbound_r=-0.2, high_eqbound_r=0.2, alpha=0.05)

```
The resulting p-value and 95% CI match the calculations by Quertemont (2011).


##*Example 4.2*

The R code from Goertzen & Cribbie, page 536, was used:

```{r}
#Running an original two t test procedure for equivalence
#######
equivint<-0.3
corxy<-0.02
n<-100
alpha<-0.05

zei <-log((1 + equivint)/(1 - equivint))/2
zcorxy <-log((1 + corxy)/(1-corxy))/2
equivt1_fz <-(zcorxy - zei)/(1/sqrt(n - 3))
pvalue1_fz <-pnorm(equivt1_fz)
equivt2_fz <-(zcorxy + zei)/(1/sqrt(n - 3))
pvalue2_fz <-1 - pnorm(equivt2_fz)
ifelse (pvalue1_fz <= alpha & pvalue2_fz <= alpha, decis_fz <-"The null hypothesis that the correlation between var1 and var2 falls outside of the equivalence interval can be rejected.", decis_fz <-"The null hypothesis that the correlation between var1 and var2 falls outside of the equivalence interval cannot be rejected.")
pvalue1_fz
pvalue2_fz
```


Which is identical to the p-values observed in the spreadsheet, and returned by the the TOSTr function:
```{r}
TOSTr(n=100, r = 0.02, low_eqbound_r=-0.3, high_eqbound_r=0.3, alpha=0.05)

```

#**Meta-analaysis based on Cohen's d**

##*Example 5.1*

We can reproduce an example for a meta-analysis from From Rogers, J. L., Howard, K. I., & Vessey, J. T. (1993). Using significance tests to evaluate equivalence between two experimental groups. Psychological Bulletin, 113(3), page 557.


For example, we obtain the following calculations for the cognitive versus behavioral contrast. Traditional z: z = ES/SE = 0.12/0.09 = 1.333, p = .091.
Traditional confidence interval: ES ? (za/2)(SE) = 0.12 ? (1.96)(0.09), or -0.056 to 0.296.
Equivalence z: 
z1 = (ES + 0.20)/SE = (0.12 + 0.20)/0.09 = 3.556, p = .000.
z2 = (ES - Q.2Q)/SE = (0.12 - 0.20)/0.09 = -0.889, p= .187.
So we table the larger p value of 0.187.
Equivalence confidence interval:
ES ? (za)(SE) = 0.12 ? (1.645)(0.09), or -0.028 to 0.268.

We can recreate this with the TOSTmeta function:
```{r}
TOSTmeta(ES=0.12, var=0.0081, se=0.09, low_eqbound_d=-0.2, high_eqbound_d=0.2, alpha=0.05)
```

#**TOST for two proportions**

##*Example 6.1*

Walker, E., & Nowacki, A. S. (2011), applies an equivalence test for the difference between two proportions to data from Staszewski, S., Keiser, P., Montaner, J., Raffi, F., Gathe, J., Brotas, V., ... & Tortell, S. (2001), comparing the response rates of two treatments for HIV. 

The following proportions were reported: prop1 = 133/262 (50.8%), prop2 = 136/265 (51.3%). The equivalence bounds were set to a proportion difference of +/- 12%. The confidence level for the TOST CI was set to 95%, so alpha was (1-TOST.CI)/2 = 0.025

Walker & Nowacki (2011) reports the following confidence intervals for the TOST: Lower = -0.09, upper = 0.08. 

We can recreate these confidence intervals with the `TOSTtwo.prop` function.

```{r}
TOSTtwo.prop(prop1 =  133/262,
             prop2 =  136/265,
             n1 = 262,
             n2 = 265, 
             low_eqbound = -0.12,
             high_eqbound = 0.12, 
             alpha = 0.025, 
             plot = TRUE)
```

##*Example 6.2*

O'reilly, R., Bishop, J., Maddox, K., Hutchinson, L., Fisman, M., & Takhar, J. (2007), compared the clinical efficiency of face-to-face- vs telepsychiatry. They tested the proportions of both the return to normal levels on a symptom inventory, and hospitalization in the first year after treatment.

For return to normal levels on symptom inventory the following proportions were reported in Table 2: prop1 = 29/148 (0.20), prop2 = 30/138 (0.22). The equivalence bounds were set to a proportion difference of +/- 15%. The confidence level for the TOST CI was set to 90%, so alpha was (1-TOST.CI)/2 = 0.05

The authors only report the lower end of the interval: Lower = -0.10

We can recreate this value with the `TOSTtwo.prop` function.

```{r}
TOSTtwo.prop(prop1 =  29/148,
             prop2 =  30/138,
             n1 = 148,
             n2 = 138, 
             low_eqbound = -0.15,
             high_eqbound = 0.15, 
             alpha = 0.05, 
             plot = TRUE)
```

For hospitalization post treatment the following proportions were reported: prop1 = 18/246 (0.073), prop2 = 15/224 (reported as 0.066, but actually 0.067). The equivalence bounds were set to a proportion difference of +/- 10%. The confidence level for the TOST CI was set to 90%, so alpha was (1-TOST.CI)/2 = 0.05. 

The authors only report the lower end of the interval: Lower = -0.03

We can recreate this value with the `TOSTtwo.prop` function.

```{r}
TOSTtwo.prop(prop1 =  18/246,
             prop2 =  15/224,
             n1 = 246,
             n2 = 224, 
             low_eqbound = -0.10,
             high_eqbound = 0.10, 
             alpha = 0.05, 
             plot = TRUE)

```


##*Example 6.3*

da Silva, Logan, & Klein, J. P. (2008)  provide and example of a study comparing peripheral blood stem cells (PBSC) versus bone marrow (BM) as a cell source for unrelated donor transplants for patients with leukemia undergoing
myeloablative transplants, using the data from Eapen et al. (2007). 

In Table 3, they report equivalence tests on BM (n=583) and BPSC (n=328) treatments, and compare treatment-related mortality (TRM). TRM was 32% (187/583) in the BM group vs. 29% (95/328) in the PBSC group. Using equivalence bounds of 10%, they report a Z-value of 2.18, p = 0.0015, 90% CI of -2.1%,8.3%).  

We can reproduce this analysis using TOSTER:

```{r}
TOSTtwo.prop(prop1 =  187/583,
             prop2 =  95/328,
             n1 = 583,
             n2 = 328, 
             low_eqbound = -0.10,
             high_eqbound = 0.10, 
             alpha = 0.05, 
             plot = TRUE)

```

By comparison, we can not reject effects of 10% or larger for overall survival (OS) in their Table 3, with 57% (331/583) vs 61% (201/328), Z = 1.62, p = 0.052, 90% CI [-10.1%;1.1%].

```{r}
TOSTtwo.prop(prop1 =  331/583,
             prop2 =  201/328,
             n1 = 583,
             n2 = 328, 
             low_eqbound = -0.10,
             high_eqbound = 0.10, 
             alpha = 0.05, 
             plot = TRUE)

```
