---
title: |
  Exploring Equivalence Testing with the Updated TOSTER R Package
shorttitle: "Updated TOSTER R Package"
author:
- name: Aaron R. Caldwell
  affil: a
  email: caldwellaaron@uams.edu
type: PREPRINT
output:
  papaja::apa6_docx: default
  rticles::tf_article:
    extra_dependencies: ["float"]
bibliography: interactcadsample.bib
#appendix: appendix.tex
abstract: |
  Despite decades of methodological guidance, researchers continue to
  misinterpret non-significant results as evidence for the absence of an effect.
  Equivalence testing via two one-sided tests (TOST) provides a principled
  alternative, but adoption remains limited by software accessibility and
  lack of training. The TOSTER R package, originally developed by Lakens (2017),
  made TOST accessible to experimental psychologists. This paper describes
  a major update to the package that expands its scope with (1) a unified
  t-test interface supporting both equivalence and minimal effect testing,
  (2) a simplified hypothesis testing interface returning standard R objects,
  (3) tools for computing and testing standardized effect sizes and correlations,
  (4) robust and resampling-based alternatives including bootstrap and
  permutation tests, (5) equivalence testing for ANOVAs, and (6) power
  analysis and sample size planning. Each capability is demonstrated with
  worked examples. TOSTER is also available as a jamovi module for
  researchers who prefer a graphical interface.
keywords: |
  statistics, bootstrap, minimal effects test, NHST, TOST, R package, robust statistics
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \hypersetup{
    colorlinks=true,
    citecolor = cyan,
    linkcolor=blue,
    filecolor=magenta,
    urlcolor=blue
    }
affiliation:
- num: a
  address: |
    Department of Biostatistics, Fay W. Boozman College of Public Health, University of Arkansas for Medical Sciences Northwest, Springdale, AR  <https://orcid.org/0000-0002-4541-6283>
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  fig.pos = "H"
)
knitr::knit_hooks$set(purl = knitr::hook_purl)
library(TOSTER)
library(ggplot2)
library(ggdist)
library(patchwork)
```

# Introduction

Researchers routinely interpret non-significant p-values as evidence that no effect exists --- a logical error that has been recognized for decades [@blandaltman95]. A non-significant result means only that the data are compatible with the null hypothesis; it does not mean the null hypothesis is true.

Equivalence testing provides a principled framework for evaluating whether an effect is small enough to be considered negligible. The two one-sided tests (TOST) procedure is the most widely used approach [@schuirmann1987]. Researchers specify an upper ($\Delta_U$) and lower ($\Delta_L$) equivalence bound based on the smallest effect size of interest (SESOI). Two one-sided tests then evaluate whether the observed effect falls within those bounds. If both tests reject at a pre-specified alpha level, the effect can be considered practically equivalent to zero [@lakens_ori].

The TOSTER R package^[All updates to the package can be found on the package's website <https://aaroncaldwell.us/TOSTERpkg>] was originally developed by @lakens_ori to introduce experimental psychologists to the concept of equivalence testing and provide an accessible implementation in R. The package provided dedicated functions for independent, paired, and one-sample t-tests, as well as correlations and meta-analytic equivalence tests. It has been widely adopted, with over 1000 citations to date.

However, the original package had a number of limitations. It offered a function-per-design interface (separate functions for each t-test type), lacked robust alternatives to parametric tests, provided no support for ANOVA equivalence testing, and had no tools for comparing effect sizes across studies. The package also did not integrate with standard R workflows and usually just printed results to the console rather than returning objects that could be manipulated or plotted.

This paper provides a tutorial-style introduction to the updated TOSTER package, organized around its major capabilities: (1) a unified t-test interface (`t_TOST` and `simple_htest`) that pairs a standard significance test with an equivalence test and supports minimal effect testing, (2) a simplified hypothesis testing interface (`simple_htest`) that returns standard `htest` objects, (3) tools for computing and testing standardized effect sizes --- including standardized mean differences, rank-based effect sizes, and correlations, (4) robust and resampling-based alternatives including the Brunner-Munzel test, bootstrap and permutation t-tests, and log-transformed methods, (5) equivalence testing for ANOVAs, and (6) power analysis and sample size planning. Each section motivates when and why a researcher would use the tool before demonstrating how.

TOSTER occupies a specific niche in the R ecosystem in that it is a frequentist toolkit for equivalence testing of standard designs. Researchers interested in Bayesian approaches to evidence for the null have excellent options (e.g., bayestestR). For equivalence-style contrasts in more complex models --- mixed models, marginal means --- the emmeans and marginaleffects packages can accommodate non-zero null hypotheses. TOSTER fills the space of accessible, self-contained equivalence testing for common, albeit simple, research designs. For a broader introduction to the methodology, see @lakens_ori, @lakens2018equivalence, @lakens2020improving, and @mazzolari2022myths. TOSTER is also available as a jamovi module for researchers who prefer a graphical interface.

# Setting Equivalence Bounds

Before applying any equivalence test, the researcher must specify the equivalence bounds --- the range of effect sizes considered too small to be meaningful. There is no single correct method for choosing these bounds; the appropriate approach depends entirely on the research question and domain.

If a researcher has an equivalence hypothesis, they should be able to articulate what magnitude of effect would be too small to matter in context. The SESOI flows from the hypothesis, not from a formula. Several general strategies exist. Anchor-based methods tie the bound to an externally meaningful criterion, such as a clinically meaningful difference or a just-noticeable difference [@sesoi]. Distribution-based methods define the bound as a fraction of the standard deviation, though this approach has limitations (discussed below in the Standardized Effect Sizes section). Cost-benefit reasoning sets the bound at the point where the cost of treating a difference as real outweighs the benefit --- for example, when additional measurement or intervention resources exceed what a small effect would justify. Researchers can draw on existing literature or normative data to identify effect magnitudes that prior work suggests are negligible. When theory is sufficiently precise, however, the bound may require no external benchmark at all, and the hypothesis itself implies what magnitude of effect would be too small to matter.

Using "default" bounds based on Cohen's benchmarks (e.g., SMD = 0.2, 0.5, 0.8 to imply small, medium, or large effects, respectively) without domain-specific justification is strongly discouraged. Cohen himself stated that these conventions were "recommended for use only when no better basis for estimating the index is available" (p. 25, as cited in @panzarella2021) and reportedly came to regret having proposed them at all [@funder2019]. The benchmarks were never intended to serve as universal thresholds, and their uncritical application can lead to equivalence bounds that are either far too liberal or far too conservative for the question at hand.

If a researcher cannot specify meaningful bounds, this itself is informative --- it suggests the equivalence hypothesis is not well enough developed to test. In such cases, it may be more productive to focus on effect estimation with confidence intervals rather than forcing a dichotomous equivalence decision. Throughout the examples in this paper, equivalence bounds are chosen for illustration; readers should not treat them as direct recommendations for their own research.

# TOST with t-tests

The `t_TOST` function provides a unified interface for equivalence and minimal effect testing with t-tests. It runs a standard nil significance test ($H_0: $) alongside the equivalence (or minimal effect) test, because in practice researchers may want both: "Is there a significant effect?" and "Can I rule out a meaningful effect?" The function handles independent, paired, and one-sample designs through the `paired` argument and accepts data via either a formula interface or raw vectors, making it a direct analog to base R's `t.test`. For researchers who only want to test a single hypothesis at a time, the `simple_htest` function (described in the next section) provides a lighter alternative.

Minimal effect testing (MET) is also supported. MET is useful when the hypothesis is that an effect *exceeds* a meaningful threshold --- the null hypothesis is equivalence, and rejection implies a practically meaningful effect (see Figure 1) [@mazzolari2022myths].

```{r hypplot, fig.width=6, fig.height=2.75, echo=FALSE, message = FALSE, warning = FALSE, fig.show='hold', fig.cap = "Type of Hypothesis"}

p1 = ggplot() +
  geom_vline(aes(xintercept = -.5),
             linetype = "dashed") +
  geom_vline(aes(xintercept = .5),
             linetype = "dashed") +
  geom_text(aes(
    y = 1,
    x = -0.5,
    vjust = -.9,
    hjust = "middle"
  ),
  angle = 90,
  label = 'Lower Bound') +
  geom_text(aes(
    y = 1,
    x = 0.5,
    vjust = 1.5,
    hjust = "middle"
  ),
  angle = 90,
  label = 'Upper Bound') +
  geom_text(aes(
    y = 1,
    x = 0,
    vjust = 1.5,
    hjust = "middle"
  ),
  #alignment = "center",
  label = "H0"
  ) +
  geom_text(aes(
    y = 1,
    x = 1.5,
    vjust = 1.5,
    hjust = "middle"
  ),
  #alignment = "center",
  label = "H1"
  ) +
  geom_text(aes(
    y = 1,
    x = -1.5,
    vjust = 1.5,
    hjust = "middle"
  ),
  #alignment = "center",
  label = "H1"
  ) +
theme_tidybayes() +
  scale_y_continuous(limits = c(0,1.75)) +
  scale_x_continuous(limits = c(-2,2)) +
  labs(x = "", y = "",
       title="Minimal Effect Test",
       caption = "H1 = Alternative Hypothesis \n H0 = Null Hypothesis") +
  theme(
    strip.text = element_text(face = "bold", size = 10),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

p2 = ggplot() +
  geom_vline(aes(xintercept = -.5),
             linetype = "dashed") +
  geom_vline(aes(xintercept = .5),
             linetype = "dashed") +
  geom_text(aes(
    y = 1,
    x = -0.5,
    vjust = -.9,
    hjust = "middle"
  ),
  angle = 90,
  label = 'Lower Bound') +
  geom_text(aes(
    y = 1,
    x = 0.5,
    vjust = 1.5,
    hjust = "middle"
  ),
  angle = 90,
  label = 'Upper Bound') +
  geom_text(aes(
    y = 1,
    x = 0,
    vjust = 1.5,
    hjust = "middle"
  ),
  #alignment = "center",
  label = "H1"
  ) +
  geom_text(aes(
    y = 1,
    x = 1.5,
    vjust = 1.5,
    hjust = "middle"
  ),
  #alignment = "center",
  label = "H0"
  ) +
  geom_text(aes(
    y = 1,
    x = -1.5,
    vjust = 1.5,
    hjust = "middle"
  ),
  #alignment = "center",
  label = "H0"
  ) +
theme_tidybayes() +
  scale_y_continuous(limits = c(0,1.75)) +
  scale_x_continuous(limits = c(-2,2)) +
  labs(x = "",
       y = "",
       title="Equivalence Test",
       caption = "H1 = Alternative Hypothesis \n H0 = Null Hypothesis") +
  theme(
    strip.text = element_text(face = "bold", size = 10),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

p1 + p2 + plot_annotation(tag_levels = 'A')
```

\newpage

## Independent Groups Example

In these examples we will use the `sleep` data and the `bugs` data from the `jmv` R package.

```{r}
data('sleep')
library(jmv)
data('bugs')
```

For the first example, we use the `sleep` data with equivalence bounds of $\pm$ 0.5 units. The formula interface works identically to `t.test`:

```{r}
res1 = t_TOST(formula = extra ~ group, data = sleep,
              eqb = .5, smd_ci = "t")
```

The `print` method provides a verbose summary including the two-tailed test, both one-sided equivalence tests, and the standardized mean difference (Hedges' $g$ by default):

```{r}
print(res1)
```

\newpage

The generic `plot` method produces a consonance density plot, inspired by the concurve R package [@rafi2020], that shows the estimate, confidence intervals, and equivalence bounds simultaneously:

```{r cdplot,fig.width=6, fig.height=5,fig.cap="Consonance density plot for the independent-samples equivalence test."}
plot(res1, type = "cd")
```

Other plot types are available: `type = "c"` produces a consonance plot showing multiple confidence interval levels, and shading can be customized with the `ci_shades` argument.

\newpage

## One-Sample t-test with Asymmetric Bounds

When the equivalence bounds are not symmetric around the null, two values can be supplied to `eqb`. This is common when the research question implies asymmetric consequences. Here we test whether the mean of `LDHF` falls within the bounds of 5.5 and 8.5 around a hypothesized value of 7.5:

```{r}
res4 = t_TOST(x = bugs$LDHF,
              hypothesis = "EQU",
              mu = 7.5,
              eqb = c(5.5,8.5))
res4
```

The same interface handles paired designs by setting `paired = TRUE`. Standardized mean differences are reported automatically; Hedges' correction is applied by default but can be toggled with `bias_correction`, and Glass's delta can be requested via the `glass` argument. Setting equivalence bounds in SMD units (via `eqbound_type = "SMD"`) is possible but discouraged because the bound becomes dependent on sample variance (see the Standardized Effect Sizes section below).

\newpage

## Using Summary Statistics

When only summary statistics are available (e.g., from a published article), the `tsum_TOST` function performs the same tests using sample means, standard deviations, sample sizes, and (for paired designs) the correlation between measures^[The `extract_r_paired` function can be used if the correlation between paired observations is not readily available.]:

```{r}
res_tsum = tsum_TOST(
  m1 = mean(bugs$LDHF, na.rm=TRUE), sd1 = sd(bugs$LDHF, na.rm=TRUE),
  n1 = length(na.omit(bugs$LDHF)),
  hypothesis = "EQU", smd_ci = "t", eqb = c(5.5, 8.5)
)

res_tsum
```

# Simplified Hypothesis Testing

The `t_TOST` function is designed to be comprehensive --- it runs three tests, computes standardized effect sizes, and produces verbose output. This is useful for exploration or situations where researchers want to pair a traditional nil significance test with an equivalence test. However, sometimes a researcher has a single specific hypothesis to test. The `simple_htest` function provides a lighter interface for this purpose. It returns a standard R `htest` object --- the same class returned by `t.test()` and `wilcox.test()` --- but with expanded `alternative` options that include `"equivalence"` and `"minimal.effect"`. This means researchers can perform equivalence testing with the same familiar interface they already use for standard hypothesis tests.

Alongside `simple_htest`, a set of helper functions makes it easy to work with any `htest` object: `as_htest` converts verbose TOST output to the simpler format, `describe_htest` generates formatted text for reporting, `df_htest` converts results to data frames for tabulation, and `plot_htest_est` produces estimate plots with equivalence bounds.

## simple_htest

The `simple_htest` function supports `t.test` and `wilcox.test` as the underlying test (via the `test` argument) and all five alternative hypotheses: `"two.sided"`, `"less"`, `"greater"`, `"equivalence"`, and `"minimal.effect"`. The `mu` parameter specifies equivalence bounds --- a single value creates symmetric bounds, or two values can be given for asymmetric bounds.

```{r}
res_sh1 <- simple_htest(extra ~ group, data = sleep,
                         alternative = "equivalence", mu = 0.5)
res_sh1
```

Notice how the output is a single clean `htest` object --- much more concise than the verbose `t_TOST` output shown previously.

## Describing and Tabulating Results

For writing up results, `describe_htest` generates a formatted text string following typical scientific reporting conventions:

```{r}
describe_htest(simple_htest(extra ~ group, data = sleep))
```


## Plotting Results

A visual summary of the estimate and confidence interval can be produced with `plot_htest_est`. For equivalence and minimal effect tests, the equivalence bounds are shown as dashed vertical lines:

```{r shplot1, fig.width=6, fig.height=3.5, fig.cap = "Estimate plot for an equivalence test via simple_htest."}
plot_htest_est(res_sh1)
```

\newpage

# Standardized Effect Sizes

This section provides a unified treatment of the standardized effect size tools in TOSTER --- standardized mean differences (SMDs), rank-based effect sizes, correlation equivalence tests, and tools for comparing SMDs between studies. Before demonstrating these tools, it is worth discussing the limitations of standardized effect sizes, because these limitations directly affect how equivalence bounds should be set.

## Problems with Standardized Effect Sizes

Standardized effect sizes --- Cohen's $d$, correlation coefficients, standardized regression coefficients --- are appealing because they allow comparison across studies and scales, are required by many journals, and serve as the primary currency of meta-analysis. However, they have well-documented problems that are especially relevant to equivalence testing.

Standardized mean differences are sensitive to the sample variance: the "same" real-world effect produces different $d$ values across populations and study designs. Cohen's $d$ conflates the magnitude of the mean difference with the precision of measurement, which means that a narrow equivalence bound on $d$ may correspond to a very different raw-scale bound depending on the variability of the sample [@Caldwell2020]. This makes it unclear what, substantively, is being declared "equivalent" --- are we claiming the means are close, or that the means are close *relative to the variance*? [@baguley2009]. The choice of denominator (pooled SD, control SD, within-subject SD) further complicates interpretation.

Correlations have analogous issues. @tukey1969 noted that correlations are influenced by the range of both variables, so the "same" bivariate relationship produces different coefficients depending on sample selection. A correlation of 0.3 in a restricted-range sample may correspond to a much larger association in the full population. Standardized regression coefficients share these vulnerabilities: they depend on the variances of both the predictor and outcome, which makes them unstable across samples and difficult to interpret substantively [@greenland1991].

The implication for equivalence testing is straightforward. When the measurement scale is meaningful, researchers should generally prefer *unstandardized* (raw-scale) equivalence bounds. When standardized bounds are used, the choice should be justified with reference to the specific population and measurement context, not borrowed from generic benchmarks.

With these caveats in mind, the following subsections describe the tools TOSTER provides for computing and testing standardized effects when they are appropriate.

## SMD Functions

The `smd_calc` function computes standardized mean differences --- Cohen's $d$, Hedges' $g$ (bias-corrected), or Glass's $\Delta$ --- with confidence intervals and optional hypothesis testing. It supports one-sample, paired, and two-sample designs, and can return either an `htest` object or a data frame. The SMD is already reported in `t_TOST` output, but `smd_calc` allows standalone computation and supports equivalence testing directly on the effect size.

```{r}
smd_calc(formula = extra ~ group, 
         data = sleep,
         paired = FALSE, 
         bias_correction = TRUE,
         alternative = "equivalence", 
         null.value = c(-0.5,0.5),
         output = "htest")
```


## Equivalence Tests for Correlations

Researchers sometimes need to test that a correlation is practically zero --- for example, when evaluating discriminant validity or confirming that a potential confounder is unrelated to an outcome. The `boot_cor_test` function supports provides robust support for testing equivalence on correlation coefficients. Bounds are specified via the `null` argument:

```{r}
X = rnorm(30)
Y = rnorm(30)

boot_cor_test(
  x = X,
  y = Y,
  method = "pearson",
  alternative = "equivalence",
  null = 0.4)
```

This extends the original TOSTER package's correlation coverage by supporting all three major correlation methods and returning standard `htest` objects that integrate with the helper functions described above. If summary statistics are only available, an approximate method is provided through `corsum_test`, which uses the Fisher z-transformation to compute confidence intervals and p-values.

## Comparing SMDs Between Studies

When evaluating whether a replication study produced results consistent with the original, researchers need a way to formally compare effect sizes. The `compare_smd` function tests whether two SMD estimates --- from an original study and a replication, for example --- differ more than would be expected by sampling variability alone.

Consider an original study reporting Cohen's $d_z$ = 0.95 with 25 participants and a replication with $d$ = 0.23 and 50 participants. Are these estimates compatible?

```{r}
compare_smd(smd1 = 0.95,
            n1 = 25,
            smd2 = 0.23,
            n2 = 50,
            paired = TRUE)
```

Setting `TOST = TRUE` with an equivalence bound via the `null` argument tests whether the two SMDs are practically equivalent:

```{r}
compare_smd(smd1 = 0.95, n1 = 25, smd2 = 0.23, n2 = 50,
            paired = TRUE, TOST = TRUE, null = .25)
```

Based on these results, we would reject the null significance hypothesis (the SMDs differ significantly) but fail to reject the equivalence null hypothesis (we cannot conclude the SMDs are practically equivalent). When raw data are available, `boot_compare_smd` provides a bootstrap alternative; see the package vignettes for details.

\newpage

**A note on argument names for equivalence bounds.** The equivalence bound argument is named differently across TOSTER functions for historical and compatibility reasons: `eqb` in `t_TOST`, `wilcox_TOST`, `equ_ftest`, and `equ_anova`; `mu` in `simple_htest`, `brunner_munzel`, `boot_t_test`, and `perm_t_test` (for compatibility with base R's `t.test`); and `null` in `compare_smd` and `z_cor_test`. Despite the different names, the interpretation is the same: the bound(s) defining the region of practical equivalence.

# Robust Methods for Equivalence Testing

In some cases, serious violations of t-test assumptions (normality, homoscedasticity) could inflate the type I error rate of TOST. TOSTER provides robust alternatives organized into three categories: rank-based tests, resampling methods, and log-transformed methods.

## Rank-Based Tests

Rank-based tests avoid distributional assumptions by operating on the ranks of the data rather than the raw values.
These tests are *not* tests of means or medians --- they target different estimands. The choice between them should be driven by the estimand of interest.

### Brunner-Munzel Test

The Brunner-Munzel test [@brunner2000] estimates the relative effect $\hat{p} = P(X > Y) + 0.5 \cdot P(X = Y)$ --- the probability that a randomly chosen observation from one group exceeds a randomly chosen observation from the other. This estimand is interpretable without distributional assumptions, unlike the WMW test which requires a location-shift assumption to say anything about location differences. @karch2021 argues it should be preferred as the default nonparametric procedure because it is valid under heteroscedasticity and provides a more interpretable effect size.

The `brunner_munzel` function supports equivalence and minimal effect testing via `alternative`, with bounds on the probability scale via `mu`. Results can be reported on different scales (`"probability"`, `"difference"`, `"logodds"`, or `"odds"`), and three test methods are available: `"t"` (default), `"logit"` (range-preserving CIs), and `"perm"` (studentized permutation; @neubert2007).

```{r}
bm_test = brunner_munzel(formula = extra ~ group,
                         data = sleep,
                         alternative = "equivalence",
                         mu = c(0.3, 0.7))
print(bm_test)
```

\newpage

### Wilcoxon-Mann-Whitney TOST

The WMW group of tests provide non-parametric tests based on ranks. Within the TOST framework, two directional location-shift tests determine if the shift is within (equivalence) or outside (minimal effect) the equivalence bounds.

However, there are important caveats. Many researchers mistakenly believe these are tests of medians, but without a location-shift assumption, the WMW test is a test of stochastic equality. Only under a location-shift assumption does it become a test of the Hodges-Lehmann pseudo-median of pairwise differences. @median_test demonstrate striking counterexamples: groups can have equal medians yet produce a significant WMW test, very different medians yet a non-significant WMW test, or even significance in the opposite direction from the median comparison.

For equivalence testing, bounds are applied to the pseudo-median, which corresponds to the mean or median difference only under the location-shift assumption. For testing hypotheses about mean differences, `perm_t_test` or `boot_t_test` are more appropriate. For stochastic superiority, `brunner_munzel` provides a clearer framework. 

```{r}
test1 = wilcox_TOST(formula = extra ~ group,
                      data = sleep,
                      paired = FALSE,
                      eqb = .5)
print(test1)
```

Based on these results, we would conclude there is no significant difference but the results are also not equivalent (i.e., an inconclusive result). But again, the estimate we are testing is the pseudo-median, not the mean or median difference, so this conclusion may not be relevant to the research question.

\newpage

## Resampling Methods

When the goal is to test hypotheses about means while relaxing distributional assumptions, resampling methods provide powerful tools. TOSTER provides bootstrap and permutation approaches, each with distinct inferential logic.

The bootstrap [@efron93] resamples with replacement to approximate the sampling distribution. P-values are tied to the CI method via CI inversion, ensuring consistency between reported CIs and p-values [@Thulin_2024]. The permutation approach shuffles group labels (two-sample) or flips signs (one-sample/paired) to construct the null distribution directly. Studentized permutation tests [@janssen1997; @chung2013] provide valid inference under heteroscedasticity and allow accurate equivalence tests [@Arboretti_Pesarin_Salmaso_2020].

Both `boot_t_test` and `perm_t_test` return standard `htest` objects and support `alternative = "equivalence"` and `alternative = "minimal.effect"`. Both also support trimmed means via the `tr` parameter for additional robustness to outliers.

### Bootstrap t-test

```{r}
set.seed(4522)
boot_t_test(formula = extra ~ group,
            data = sleep,
            alternative = "equivalence",
            mu = c(-0.5, 0.5),
            R = 999)
```

### Permutation t-test

When `R = NULL`, the function enumerates all possible permutations for an exact test; otherwise, it draws `R` random permutations with the "+1" correction [@phipson2010]. By default the test is studentized (`perm_se = TRUE`).

```{r}
set.seed(8812)
perm_t_test(formula = extra ~ group,
            data = sleep,
            alternative = "equivalence",
            mu = c(-0.5, 0.5),
            R = 999)
```

### Bootstrap TOST

For situations where both a nil hypothesis test and an equivalence test are desired simultaneously with bootstrapping, `boot_t_TOST` returns a `TOSTt` object with bootstrapped SMD estimates [@Kirby2013]:

```{r}
set.seed(891111)
test1 = boot_t_TOST(formula = extra ~ group,
                    data = sleep,
                    paired = TRUE,
                    eqb = .5,
                    R = 999)
print(test1)
```

\newpage

## Log-Transformed Methods

The natural log transformation stabilizes variance and approximates normality for right-skewed data [@logtest]. An additional advantage is that differences on the log scale back-transform to *ratios*, making interpretation straightforward: if two samples have geometric means of 7 and 10.5, the ratio is 1.5 (i.e., 10.5/7). This property is central to pharmaceutical bioequivalence, where the FDA requires comparisons on the log-transformed scale with equivalence bounds of 0.80 to 1.25 for the ratio of geometric means [@fda; @he2022].

An alternative to `log_TOST` is to log-transform data manually and analyze with `t_TOST` or the resampling methods. Differences on the log scale can then be interpreted as "sympercents" --- symmetric percentage differences on the $100 \cdot \ln$ scale [@cole2000] --- which have the attractive property that +10 and -10 sympercents are of equal magnitude, unlike ordinary percentages.

### Log TOST

The `log_TOST` function accepts paired and two-sample comparisons (one-sample tests are not supported since there is no ratio to calculate). Instead of an SMD, a ratio of means is reported [@lajeunesse2015bias]. The default equivalence bounds are set to the FDA standard (`eqb = 1.25`; the reciprocal is taken as the other bound).

```{r, error=FALSE}
log_TOST(mpg ~ am, data = mtcars)
```

\newpage

### Bootstrap Log TOST

The bootstrap version uses the same bootstrap methods as `boot_t_test` on the log-transformed values:

```{r, error=FALSE}
boot_log_TOST(mpg ~ am, data = mtcars, R=999)
```

From this analysis, we would conclude there is a significant effect that is not practically equivalent.

\newpage

# Equivalence Testing with ANOVAs

Many researchers use ANOVA as an omnibus test before inspecting pairwise comparisons, particularly in factorial designs. As @Campbell_2021 note, the lack of a significant ANOVA result does not indicate that a factor has no effect. Equivalence testing for ANOVAs addresses this directly by asking: can we reject the hypothesis that the partial $\eta^2$ exceeds a specified bound?

The equivalence test uses the non-central *F* distribution. The non-centrality parameter is computed from the equivalence bound and the degrees of freedom, and the *p*-value is obtained from the cumulative distribution function of the non-central *F*. I have extended the approach of @Campbell_2021 --- who proposed equivalence testing for one-way ANOVAs --- to accommodate multi-factor and factorial designs by generalizing the non-centrality parameter calculation. The full derivation is described in the package documentation; here we demonstrate usage.

## Example with Summary Statistics

Using the `InsectSprays` data and the `equ_ftest` function, which accepts summary *F*-test statistics directly:

```{r warning=FALSE, message=FALSE}
data("InsectSprays")
aovtest = aov(count ~ spray, data = InsectSprays)
anova(aovtest)

```

```{r}
equ_ftest(Fstat = 34.70228,  df1 = 5, df2 = 66,  eqb = 0.35)
```

Based on these results, there is a significant effect of "spray" and the differences are *not* statistically equivalent --- we reject the traditional null hypothesis but fail to reject the equivalence null hypothesis.

\newpage

## Example with R ANOVA Objects

The `equ_anova` function accepts objects from `stats::aov`, `car::Anova`, and `afex::aov_car`, avoiding the need to extract summary statistics manually:

```{r}
# Example using a purely within-subjects design
# (Maxwell & Delaney, 2004, Chapter 12, Table 12.5, p. 578):
library(afex)
data(md_12.1)
aovtest2 = aov_ez("id", "rt", md_12.1, within = c("angle", "noise"),
       anova_table=list(correction = "none", es = "none"))
equ_anova(aovtest2,
          eqb = 0.35)
```

\newpage

# Power Analysis and Sample Size Planning

Researchers planning an equivalence study will immediately need to determine the required sample size. Power for equivalence tests behaves differently than for standard tests: larger samples are generally needed, and power depends on both the true effect size and the width of the equivalence bounds.

## Power for TOST t-tests

The `power_t_TOST` function computes exact power for TOST with one-sample, two-sample, and paired t-tests. It can solve for any one of `n`, `power`, or `alpha` given the others. The `delta` argument specifies the true mean difference (set to 0 if the true effect is exactly zero) and `eqb` specifies the equivalence bounds.

For example, to determine the sample size needed to achieve 80% power to detect equivalence within bounds of $\pm$ 0.5 (assuming the true difference is zero and $SD$ = 1):

```{r}
power_t_TOST(delta = 0, sd = 1, eqb = 0.5,
             alpha = 0.05, power = 0.8,
             type = "two.sample")
```

## Power for Correlations

The `power_z_cor` function provides power analysis for equivalence tests on Pearson correlations using Fisher's z-transformation:

```{r}
power_z_cor(rho = 0, power = 0.8,
            null = 0.3, alpha = 0.05,
            alternative = "equivalence")
```

## Extensions: PowerTOST and Superpower

For more complex designs --- crossover studies, replicate designs, group-sequential equivalence trials --- the PowerTOST package [@PowerTOST] provides extensive functionality beyond what TOSTER offers. For factorial designs where equivalence contrasts are tested via estimated marginal means, simulation-based power analysis via the Superpower package is recommended. Both packages complement TOSTER's built-in power tools for standard designs.

\newpage

# Conclusions

This paper has described the major updates to the TOSTER R package, expanding its scope from the original t-test and correlation equivalence tests to a comprehensive toolkit for frequentist equivalence testing. The update adds a unified t-test interface with minimal effect testing support, a simplified hypothesis testing interface that integrates with the standard R ecosystem, tools for standardized effect sizes and correlations, robust and resampling-based alternatives, ANOVA equivalence testing, and power analysis for sample size planning.

Several limitations should be acknowledged. TOSTER is designed for standard research designs --- it does not handle multilevel or mixed-effects models, latent variable models, or survival data. For these settings, researchers should consider packages like emmeans or marginaleffects, which can accommodate non-zero null hypotheses in more complex modeling frameworks. TOSTER also does not provide Bayesian equivalence testing; researchers interested in Bayes factors for the null should consult packages such as BayesFactor.

I hope that the updated package continues to build upon the original impact of @lakens_ori and makes equivalence testing more accessible to researchers across disciplines. At the very least, I would encourage users to compare results across methods (e.g., parametric vs. robust) to ensure that conclusions do not depend on the chosen analysis. If the interpretation substantially changes depending on the test, this suggests features of the data --- violations of assumptions, outliers, distributional shape --- that warrant further investigation.

\newpage

# Additional Information

The TOSTER package can be installed from CRAN or from GitHub:

```
# Install from CRAN
install.packages("TOSTER")

# Install the development version from GitHub
devtools::install_github("Lakens/TOSTER")
```

## Acknowledgement(s) {-}

I would like to thank everyone from the Lakens laboratory group for their input and suggestions.

## Disclosure statement {-}

The author of this manuscript is the author of the TOSTER package.
Citations of this manuscript will benefit his citation count.

## Funding {-}

No funding was provided for this work.

## Notes on contributor(s) {-}

Daniel Lakens provided a review of many of the materials that have been incorporated
into the update of TOSTER, and was the original author of this package.
Without his help and encouragement, the TOSTER package and this update would not exist.

## Nomenclature/Notation {-}

- ANOVA: Analysis of Variance
- Bootstrapping: the use of random sampling with replacement to estimate statistics
- FDA: Food and Drug Administration (United States of America)
- MET: Minimal Effects Test
- ncp: non-centrality parameter
- SESOI: Smallest Effect Size of Interest
- SMD: Standardized Mean Difference (e.g., Cohen's d)
- TOST: Two One-Sided Tests
- WMW: Wilcoxon-Mann-Whitney

## Notes {-}

The R package is also (partially) implemented in jamovi as the TOSTER module.

\newpage

# References
