---
title: |
  Exploring Equivalence Testing with the Updated TOSTER R Package
shorttitle: "Updated TOSTER R Package"
author:
- name: Aaron R. Caldwell
  affil: a
  email: caldwellaaron@uams.edu
type: PREPRINT
output:
  rticles::tf_article:
    extra_dependencies: ["float"]
  papaja::apa6_docx: default
bibliography: interactcadsample.bib
#appendix: appendix.tex
abstract: |
  Despite decades of methodological guidance, researchers continue to
  misinterpret non-significant results as evidence for the absence of an effect.
  Equivalence testing via two one-sided tests (TOST) provides a principled
  alternative, but adoption remains limited by software accessibility and
  lack of training. The TOSTER R package, originally developed by Lakens (2017),
  made TOST accessible to experimental psychologists. This paper describes
  a major update to the package that expands its scope with (1) a unified
  t-test interface supporting both equivalence and minimal effect testing,
  (2) a simplified hypothesis testing interface returning standard R objects,
  (3) tools for computing and testing standardized effect sizes and correlations,
  (4) robust and resampling-based alternatives including bootstrap and
  permutation tests, (5) equivalence testing for ANOVAs, and (6) power
  analysis and sample size planning. Each capability is demonstrated with
  worked examples. TOSTER is also available as a jamovi module for
  researchers who prefer a graphical interface.
keywords: |
  statistics, bootstrap, minimal effects test, NHST, TOST, R package, robust statistics
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \providecommand{\pandocbounded}[1]{#1}
  \hypersetup{
    colorlinks=true,
    citecolor = cyan,
    linkcolor=blue,
    filecolor=magenta,
    urlcolor=blue
    }
affiliation:
- num: a
  address: |
    Department of Biostatistics, Fay W. Boozman College of Public Health, University of Arkansas for Medical Sciences Northwest, Springdale, AR  <https://orcid.org/0000-0002-4541-6283>
editor_options: 
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  fig.pos = "H",
  fig.width = 6,
  fig.height = 4
)
knitr::knit_hooks$set(purl = knitr::hook_purl)
library(TOSTER)
library(ggplot2)
library(ggdist)
library(patchwork)
```

# Introduction

Researchers routinely interpret non-significant p-values from a nil-hypothesis (i.e., null hypothesis that the mean difference equals zero) test as evidence that no effect exists --- a logical error that has been recognized for decades [@blandaltman95]. A non-significant result means only that the data are compatible with the null hypothesis; it does not mean the null hypothesis is true.

Equivalence testing provides a principled framework for evaluating whether an effect is small enough to be considered negligible. The two one-sided tests (TOST) procedure is the most widely used approach [@schuirmann1987]. Researchers specify an upper ($\Delta_U$) and lower ($\Delta_L$) equivalence bound based on the smallest effect size of interest (SESOI). Two one-sided tests then evaluate whether the observed effect falls within those bounds. If both tests reject at a pre-specified alpha level, the effect can be considered practically equivalent to zero [@lakens_ori].

The TOSTER R package^[All updates to the package can be found on the package's website <https://aaroncaldwell.us/TOSTERpkg>] was originally developed by @lakens_ori to introduce experimental psychologists to the concept of equivalence testing and provide an accessible implementation in R. The package provided dedicated functions for independent, paired, and one-sample t-tests, as well as correlations and meta-analytic equivalence tests. It has been widely adopted, with over 1000 citations to date.

However, the original package had a number of limitations. It offered a function-per-design interface (separate functions for each t-test type), lacked robust alternatives to parametric tests, provided no support for ANOVA equivalence testing, and had no tools for comparing effect sizes across studies. The package also did not integrate with standard R workflows and usually just printed results to the console rather than returning objects that could be manipulated or plotted.

This paper provides a tutorial-style introduction to the updated TOSTER package, organized around its major capabilities: (1) a unified t-test interface (`t_TOST`) that pairs a standard significance test with an equivalence test and supports minimal effect testing, (2) a simplified hypothesis testing interface (`simple_htest`) that returns standard `htest` objects, (3) tools for computing and testing standardized effect sizes --- including standardized mean differences (SMDs), rank-based effect sizes, and correlations, (4) robust and resampling-based alternatives including the Brunner-Munzel test, bootstrap and permutation t-tests, and log-transformed methods, (5) equivalence testing for ANOVAs, and (6) power analysis and sample size planning. Each section motivates when and why a researcher would use the tool before demonstrating how.

TOSTER occupies a specific niche in the R ecosystem in that it is a frequentist toolkit for equivalence testing of standard designs. Researchers interested in Bayesian approaches to evidence for the null have excellent options (e.g., `bayestestR`). For equivalence-style contrasts in more complex models --- mixed models, marginal means --- the `emmeans` and `marginaleffects` packages can accommodate non-zero null hypotheses. TOSTER fills the space of accessible, self-contained equivalence testing for common, albeit simple, research designs. For a broader introduction to the methodology, see @lakens_ori, @lakens2018equivalence, @lakens2020improving, @mazzolari2022myths, and @Tomek_Caldwell_Eisner_2026. TOSTER is also available as a [jamovi](https://www.jamovi.org/) module for researchers who prefer a graphical interface although it has limited capabilities compared to the R package.

# Setting Equivalence Bounds

Two decisions precede any equivalence test: *what* quantity the hypothesis concerns, and *how much* of that quantity is too small to matter. The first is a choice of estimand, the second a choice of bounds. The two are not separable, because a bound is a magnitude on the scale of some particular quantity, and the same number can describe quite different hypotheses depending on what is being estimated.

## Choosing an Estimand

The estimand is the quantity a study is attempting to estimate [@Lundberg_Johnson_Stewart_2021]. In an ideal scenario, the hypothesis should determine the estimand, and estimand then narrows the set of appropriate tests. Inheriting an estimand from whichever procedure happens to be run is a reliable way to end up testing the wrong hypothesis.

This matters more for equivalence testing than for nil-hypothesis testing. When two distributions are identical, the location-type quantities in Table \ref{tab:estimands} all take their null values at once --- the mean difference and the pseudo-median are both zero, and the probability of superiority is exactly 0.5 --- so a nil test is comparatively forgiving of an unexamined estimand. An equivalence bound makes the decision more consequential. It is a specific magnitude on a specific scale, and there is generally no way to translate a bound on one scale into a bound on another (at least without making very serious assumptions). Bounds of $\pm$ 0.5 placed on a mean difference, on a pseudo-median, and on the SMD describe three different hypotheses, only one of which is likely to be the intended one.

Table \ref{tab:estimands} lists the estimands TOSTER can place bounds on and the functions that target each. Two entries deserve comment. The trimmed mean, available through the `tr` argument of the resampling functions (`tr = 0` by default), coincides with the mean only when the distribution is symmetric; setting `tr` therefore changes the estimand rather than merely making the same estimate more robust. The SMD is a rescaling of the mean difference by a sample-dependent quantity, which is why it warrants the separate caution given later in this paper.

Table: \label{tab:estimands}Estimands supported by TOSTER, the scale on which equivalence bounds are specified for each, and the primary functions targeting them.

| Estimand | Scale of the bound | Primary functions |
|:---------------------------|:--------------------|:--------------------------|
| Mean or mean difference | Original units | `t_TOST`, `tsum_TOST`, `simple_htest` |
| Mean, with empirically calibrated reference distributions | Original units | `boot_t_TOST`, `boot_t_test`, `perm_t_test` |
| Trimmed mean difference | Original units | `boot_t_TOST`, `perm_t_test` (via `tr`) |
| Ratio of geometric means | Multiplicative ratio | `log_TOST`, `boot_log_TOST` |
| Pseudo-median (Hodges-Lehmann) | Original units | `wilcox_TOST` |
| Stochastic superiority | Probability (0 to 1) | `brunner_munzel` |
| Standardized mean difference | Standard deviation units | `smd_calc`, `boot_smd_calc` |
| Correlation | Correlation ($-1$ to 1) | `z_cor_test`, `boot_cor_test` |
| Proportion of variance explained | Proportion (0 to 1) | `equ_anova`, `equ_ftest` |

In applied settings the estimand is not fully pinned down by the choice of summary statistic either. Decisions about how intercurrent events --- post-randomization events such as dropout, treatment discontinuation, or the use of rescue medication --- and missing data are handled determine which population summary is being estimated at all. Two analyses that both report "the mean difference" may be estimating materially different things if they resolve these questions differently.

The practical upshot is to settle the estimand first and then justify a bound on its scale, rather than selecting a test and adopting whatever quantity it happens to return. This ordering also makes the robust methods described later in this paper easier to navigate: they are not a menu of interchangeable fixes for assumption violations, but tests of different quantities, and only some of them remain answers to a hypothesis about means.

## Justifying the Bounds

Before applying any equivalence test, the researcher must specify the equivalence bounds which defines the range of effect sizes considered too small to be meaningful. There is no single correct method for choosing these bounds; the appropriate approach depends entirely on the research question and domain of research.

If a researcher has an equivalence hypothesis, they should be able to articulate what magnitude of effect would be too small to matter in context. The SESOI must flow from the hypothesis formulated by theory and context not from a formula. However, there are several general strategies that have been evaluated [@delta2]. Anchor-based methods tie the bound to an externally meaningful criterion, such as a clinically meaningful difference or a "just-noticeable" difference [@sesoi]. Distribution-based methods define the bound as a fraction of the measurement error [@Jacobson_Truax_1991], though this approach is generally not recommended [@delta2]. Cost-benefit, also referred to as health economic methods [@delta2], reasoning sets the bound at the point where there is a net benefit relative to the cost of an intervention. Researchers can draw on existing literature or normative data to identify effect magnitudes that prior work suggests are negligible. When theory is sufficiently precise, however, the bound may require no external benchmark at all, and the hypothesis itself implies what magnitude of effect would be too small to matter.

Using "default" bounds on the SMD (also known as Cohen's d) based on Cohen's benchmarks (e.g., SMD = 0.2, 0.5, 0.8 to imply small, medium, or large effects, respectively) without domain-specific justification is strongly discouraged. Cohen himself stated that these conventions were "recommended for use only when no better basis for estimating the index is available" (p. 25, as cited in @panzarella2021) and reportedly came to regret having proposed them at all [@funder2019]. These benchmarks were never intended as universal thresholds; applied uncritically, they can produce equivalence bounds that are too wide, too narrow, and likely entirely disconnected from the substantive research question.

If a researcher cannot specify meaningful bounds, this itself is informative as it suggests the equivalence hypothesis is not well enough developed to test such a hypothesis. In such cases, it may be more productive to focus on effect estimation and description rather than leaping to hypothesis tests [@scheel_hypothesis]. Throughout the examples in this paper, equivalence bounds, and the data sets themselves, are chosen for illustrative purposes; readers should not treat them as direct recommendations for their own research.

# TOST with t-tests

The t-test is the standard tool for comparing means. Its construction is simple: we estimate a mean (one-sample), the mean of the paired differences (paired samples), or the difference in means between two groups, and then divide that estimate by its standard error. The resulting t-statistic is compared against a t-distribution to obtain p-values and confidence intervals. Written generally, the statistic takes the form $t = (\hat{\mu} - \mu_0)/SE_{\hat{\mu}}$, where $\hat{\mu}$ is the estimate and $\mu_0$ is the value being tested against.

What matters for equivalence testing is that $\mu_0$ is a choice rather than a fixed feature of the test. A traditional nil hypothesis test fixes $\mu_0 = 0$ and asks whether the mean, or mean difference, is exactly zero. TOST changes none of the underlying machinery; it changes only what is substituted for $\mu_0$. Two one-sided tests are performed --- one against the lower bound ($\mu_0 = \Delta_L$) and one against the upper bound ($\mu_0 = \Delta_U$) --- and equivalence is concluded only if both reject. Because each test is one-sided at $\alpha$, the interval estimate corresponding to the procedure is a $1-2\alpha$ (typically 90%) confidence interval, which is what TOSTER reports and plots.

Throughout, the estimand for a t-test is the mean, or mean difference, on the original scale of measurement. As discussed above, this is a default rather than an inviolable property of the test, and it is worth restating here because several of the alternatives described later in this paper --- the rank-based tests in particular --- estimate different quantities entirely, and are not interchangeable substitutes when the hypothesis concerns means.

The `t_TOST` function in TOSTER provides a unified interface for equivalence and minimal effect testing with t-tests. It runs a standard nil significance test alongside the equivalence (or minimal effect) test, because in practice researchers may want both: "Is there a significant effect?" and "Can I rule out a meaningful effect?" The function handles independent, paired, and one-sample designs through the `paired` argument and accepts data via either a formula interface or raw vectors, making it a direct analog to base R's `t.test`. Additionally, the output includes an SMD calculation alongside the unstandarized mean difference with appropriate, albeit approximate, confidence intervals. For researchers who only want to test a single hypothesis at a time, the `simple_htest` function (described in the next section) provides a lighter alternative.

Minimal effect testing (MET) is also supported. MET is useful when the hypothesis is that an effect *exceeds* a meaningful threshold --- the null hypothesis is equivalence, and rejection implies a practically meaningful effect (see Figure 1) [@mazzolari2022myths].

```{r hypplot, fig.width=6, fig.height=2.75, echo=FALSE, message = FALSE, warning = FALSE, fig.show='hold', fig.cap = "Types of hypotheses tests supported by two one-sided tests (TOST) procedures."}

p1 = ggplot() +
  geom_vline(aes(xintercept = -.5),
             linetype = "dashed") +
  geom_vline(aes(xintercept = .5),
             linetype = "dashed") +
  geom_text(aes(
    y = 1,
    x = -0.5,
    vjust = -.9,
    hjust = "middle"
  ),
  angle = 90,
  label = 'Lower Bound') +
  geom_text(aes(
    y = 1,
    x = 0.5,
    vjust = 1.5,
    hjust = "middle"
  ),
  angle = 90,
  label = 'Upper Bound') +
  geom_text(aes(
    y = 1,
    x = 0,
    vjust = 1.5,
    hjust = "middle"
  ),
  #alignment = "center",
  label = "H0"
  ) +
  geom_text(aes(
    y = 1,
    x = 1.5,
    vjust = 1.5,
    hjust = "middle"
  ),
  #alignment = "center",
  label = "H1"
  ) +
  geom_text(aes(
    y = 1,
    x = -1.5,
    vjust = 1.5,
    hjust = "middle"
  ),
  #alignment = "center",
  label = "H1"
  ) +
theme_tidybayes() +
  scale_y_continuous(limits = c(0,1.75)) +
  scale_x_continuous(limits = c(-2,2)) +
  labs(x = "", y = "",
       title="Minimal Effect Test",
       caption = "H1 = Alternative Hypothesis \n H0 = Null Hypothesis") +
  theme(
    strip.text = element_text(face = "bold", size = 10),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

p2 = ggplot() +
  geom_vline(aes(xintercept = -.5),
             linetype = "dashed") +
  geom_vline(aes(xintercept = .5),
             linetype = "dashed") +
  geom_text(aes(
    y = 1,
    x = -0.5,
    vjust = -.9,
    hjust = "middle"
  ),
  angle = 90,
  label = 'Lower Bound') +
  geom_text(aes(
    y = 1,
    x = 0.5,
    vjust = 1.5,
    hjust = "middle"
  ),
  angle = 90,
  label = 'Upper Bound') +
  geom_text(aes(
    y = 1,
    x = 0,
    vjust = 1.5,
    hjust = "middle"
  ),
  #alignment = "center",
  label = "H1"
  ) +
  geom_text(aes(
    y = 1,
    x = 1.5,
    vjust = 1.5,
    hjust = "middle"
  ),
  #alignment = "center",
  label = "H0"
  ) +
  geom_text(aes(
    y = 1,
    x = -1.5,
    vjust = 1.5,
    hjust = "middle"
  ),
  #alignment = "center",
  label = "H0"
  ) +
theme_tidybayes() +
  scale_y_continuous(limits = c(0,1.75)) +
  scale_x_continuous(limits = c(-2,2)) +
  labs(x = "",
       y = "",
       title="Equivalence Test",
       caption = "H1 = Alternative Hypothesis \n H0 = Null Hypothesis") +
  theme(
    strip.text = element_text(face = "bold", size = 10),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

p1 + p2 + plot_annotation(tag_levels = 'A')
```

\newpage

## Independent Groups Example

In these examples we will use the `sleep` data and the `bugs` data from the `jmv` R package.

```{r}
data('sleep')
library(jmv)
data('bugs')
```

For the first example, we use the `sleep` data with equivalence bounds (set by the `eqb` argument) of $\pm$ 0.5 units. The `smd_ci` argument can also be set to use one of four possible calculations for the SMD confidence interval. In the example below, I have opted to go with the normal approximation for computational ease; such approximations tend provide adequate coverage [@Viechtbauer_2007]. The formula interface works almost identically to `t.test`:

```{r}
res1 = t_TOST(
  # difference in "group" of outcome "extra"
  formula = extra ~ group,
  # data
  data = sleep,
  # equivalence bound
  eqb = .5,
  # sets type of SMD confidence interval
  smd_ci = "z"
)
```

The `print` method provides a verbose summary including the two-tailed test, both one-sided equivalence tests, and the SMD (Hedges' $g$ by default):

```{r}
print(res1)
```

A greater description of the results can be accomplished with `describe` method.

```{r, results = "asis"}
cat(describe(res1))
```

\newpage

The default visualization provided by the generic `plot` method for `t_TOST`
provides a simple visualization of both effect sizes and their respective confidence intervals alongside the equivalence bounds. Test information is also printed at the top of the plot.

```{r genericplot,fig.width=6, fig.height=5,fig.cap="Results of the independent-samples equivalence test visualized with the default plot method. The top panel displays the raw mean difference with its 90% confidence interval, and the bottom panel displays the standardized effect size (Hedges's g~av~) with its 90% confidence interval. In both panels, the dashed vertical lines indicate the equivalence bounds (raw: -0.5 to 0.5; standardized: approximately -0.3 to 0.3), and the point estimate is marked by a filled circle. A 90% confidence interval is used because it corresponds to the two one-sided tests (TOST) procedure at an alpha level of 0.05."}
plot(res1)
```

Also when the generic `plot` method has the `type` argument set to "cd" this produces a consonance density plot, inspired by the "concurve" R package [@rafi2020], that shows the estimate, confidence intervals, and equivalence bounds simultaneously:

```{r cdplot,fig.width=6, fig.height=5,fig.cap="Consonance density plot for the independent-samples equivalence test. The top panel displays the standardized effect size (Hedges's g~av~) and the bottom panel displays the raw mean difference. In each panel, the consonance density curve represents the distribution of confidence intervals across all levels, with color-coded regions corresponding to the 68%, 90%, 95%, and 99.9% confidence intervals. The point estimate is marked by a filled circle along the x-axis. The dashed vertical lines indicate the equivalence bounds (standardized: -0.26 to 0.26; raw: -0.5 to 0.5). This visualization conveys the full range of effect sizes compatible with the data at varying confidence levels, rather than relying on a single confidence interval. Wider, less certain intervals (e.g., 99.9%) extend further from the point estimate, while narrower intervals (e.g., 68%) cluster near it."}
plot(res1, type = "cd")
```

Other plot types are available: `type = "c"` produces a consonance plot showing multiple confidence interval levels, and shading can be customized with the `ci_shades` argument.

\newpage

## One-Sample t-test with Asymmetric Bounds

When the equivalence bounds are not symmetric around the null, two values can be supplied to `eqb`. This is common when the research question implies asymmetric consequences. Similarly, we may want to test the mean itself rather than a mean difference. This harkens back to the original use of the t-test by Gosset at Guinness Brewery, where the concern was whether a single sample mean fell within acceptable limits. Here we test whether the mean of LDHF falls within the bounds of 5.5 and 8.5 around a hypothesized value of 7.5:

```{r}
res4 = t_TOST(
  # single sample, vector from which to estimate mean
  x = bugs$LDHF,
  # set nil significance test to 7.5
  mu = 7.5,
  # set TOST bounds to 5.5 and 8.5
  eqb = c(5.5, 8.5)
)
res4
```

The same interface handles paired designs by setting `paired = TRUE`. SMDs are reported automatically; Hedges's correction is applied by default but can be toggled with `bias_correction`, and Glass's delta can be requested via the `glass` argument. Setting equivalence bounds in SMD units (via `eqbound_type = "SMD"`) is possible but discouraged because the bound becomes dependent on sample variance (see the Standardized Effect Sizes section below).

## Checking Assumptions 

The parametric t-tests implemented in `t_TOST` assume independent observations, approximately normal residuals (or differences, in the paired and one-sample cases), and, for Student's form, equal variances between groups; Welch's correction relaxes the equal-variance assumption and is the default. The residual normality assumption matters most in small samples, where the central limit theorem provides limited protection. Visual diagnostics are generally more informative than formal tests: QQ plots of residuals or differences for normality, side-by-side boxplots or density overlays for variance comparison, and inspection of residual plots or boxplots for extreme or influential observations. A worked diagnostic workflow for each design is provided in the online supplement. When assumptions are seriously in doubt, the bootstrap and permutation alternatives discussed in later sections provide valid inference under relaxed --— though not eliminated —-- distributional requirements.

\newpage

## Using Summary Statistics

When only summary statistics are available (e.g., from a published article), the `tsum_TOST` function performs the same tests using sample means, standard deviations, sample sizes, and (for paired designs) the correlation between measures^[The `extract_r_paired` function can be used if the correlation between paired observations is not readily available.]:

```{r}
res_tsum = tsum_TOST(
  # sample mean
  m1 = mean(bugs$LDHF, na.rm = TRUE),
  # sample standard deviation
  sd1 = sd(bugs$LDHF, na.rm = TRUE),
  # sample size
  n1 = length(na.omit(bugs$LDHF)),
  # equivalence rather than minimal effects test
  hypothesis = "EQU",
  # sets type of SMD confidence interval
  smd_ci = "t",
  # equivalence bounds
  eqb = c(5.5, 8.5)
)

res_tsum
```

# Simplified Hypothesis Testing

The `t_TOST` function is designed to be comprehensive in that it runs three tests, computes standardized effect sizes, and produces verbose output. This is useful for exploration or situations where researchers *always* want to pair a traditional nil significance test with an equivalence test. However, sometimes a researcher has a single specific hypothesis to test, or simply wants less verbose output (e.g., SMD calculation is not wanted). The `simple_htest` function provides a lighter interface for this purpose. It returns a standard R `htest` object --- the same class returned by `t.test()` and `wilcox.test()` --- but with expanded `alternative` options that include `"equivalence"` and `"minimal.effect"`. This means researchers can perform equivalence testing with the same familiar interface they already use for standard hypothesis tests^[In fact, users can mostly replace `t.test()` and `wilcox.test()` with `simple_htest` as it also allows for the standard alternative hypotheses of two-sided, greater, less. If you have a hypothesis that would traditionally utilize a t-test or Wilcoxon-Mann-Whitney (WMW) test, then you can jump right to using `simple_htest` in TOSTER].

Alongside `simple_htest`, a set of helper functions makes it easy to work with any `htest` object: `as_htest` converts verbose TOST output to the simpler format, `describe_htest` generates formatted text for reporting, `df_htest` converts results to data frames for tabulation, and `plot_htest_est` produces estimate plots with equivalence bounds. Checking assumptions would follow the same workflow as `t_TOST`, as both are using t-tests, as discussed in the online supplement.

## The simplified function

The `simple_htest` function supports `t.test` and `wilcox.test` as the underlying test (via the `test` argument) and all five alternative hypotheses: `"two.sided"`, `"less"`, `"greater"`, `"equivalence"`, and `"minimal.effect"`. The `mu` parameter specifies equivalence bounds --- a single value creates symmetric bounds (e.g., if 1 is provided at `mu` with `alternative = "equivalence"` then the equivalence region will be [-1,1]), or two values can be given for asymmetric bounds.

```{r}
res_sh1 = simple_htest(
  extra ~ group,
  data = sleep,
  paired = TRUE,
  # equivalence hypothesis
  alternative = "equivalence",
  # symmetric bounds of -0.5 and 0.5
  mu = 0.5
)
res_sh1
```

Additionally, tests other that TOST can be applied with this function. The traditional nil-hypothesis tests, non-inferiority, or even superiority-by-a-margin hypotheses can all be tested with this function. For example, let's say we want to test that a treatment group in the sleep data was just non-inferior to the other treatment rather than equivalent. This can be accomplished by just setting the direction with the `alternative` argument and the margin of non-inferiority by the `mu` argument.

```{r}
res_sh2 = simple_htest(
  extra ~ group,
  data = sleep,
  paired = TRUE,
  # non-inferiority rather than equivalence
  alternative = "greater",
  # margin of non-inferiority
  mu = -0.5
)
res_sh2
```

Notice how the output, in either hypothesis test above, is a single clean `htest` object. This is much more concise than the verbose `t_TOST` output shown previously on purpose. If you want to perform additional tests --- such as a complementary nil significance tests --- then that has to be done separately. However, there are some methods (described below) that make this `htest` format easy to deal with when reporting statistical test results.

## Describing and Tabulating Results

For writing up results, `describe_htest` generates a formatted text string following typical scientific reporting conventions:

```{r, results = "asis"}
cat(describe_htest(res_sh1))
```


## Plotting Results

A visual summary of the estimate and confidence interval can be produced with `plot_htest_est`. For equivalence and minimal effect tests, the equivalence bounds are shown as dashed vertical lines:

```{r shplot1, fig.width=6, fig.height=3.5, fig.cap = "Estimate plot for an equivalence test via simple_htest."}
plot_htest_est(res_sh1)
```

\newpage

# Standardized Effect Sizes

This section provides a unified treatment of the standardized effect size tools in TOSTER and I cover SMDs, correlation coefficients, and tools for comparing both between studies. Before demonstrating these tools, it is worth discussing the limitations of standardized effect sizes, because these limitations are especially important for equivalence testing.

## A Brief Warning on Standardized Effect Sizes

Standardized effect sizes --- Cohen's $d$, correlation coefficients, standardized regression coefficients --- are appealing because they allow comparison across studies and scales, are required by many journals, and serve as the primary currency of meta-analysis. However, they have well-documented problems that are especially relevant to equivalence testing.

SMDs are sensitive to the sample variance. The "same" real-world effect produces different $d$ values across populations and study designs. The ubiquitously reported Cohen's $d$ conflates the magnitude of the mean difference with the precision of measurement, which means that a narrow equivalence bound on $d$ may correspond to a very different raw-scale bound depending on the variability of the sample [@Caldwell2020; @baguley2009]. This makes it unclear what, substantively, is being declared "equivalent" --- are we claiming the means are close, or that the means are close *relative to the variance*? The choice of denominator (pooled SD, control SD, within-subject SD) further complicates that same dubious interpretation.

Correlations have analogous issues. @tukey1969 noted that correlations are influenced by the range of both variables, so the "same" bivariate relationship produces different coefficients depending on sample selection. A correlation of 0.3 in a restricted-range sample may correspond to a much larger association in the full population. Standardized regression coefficients share these vulnerabilities. These depend on the variances of both the predictor and outcome, which makes them unstable across samples and difficult to interpret substantively [@greenland1991].

The implication for equivalence testing is straightforward. When the measurement scale is meaningful, researchers should generally prefer *unstandardized* (raw-scale) equivalence bounds. When standardized bounds are used, the choice should be justified with reference to the specific population and measurement context, not borrowed from generic benchmarks.

With these caveats in mind, the following subsections describe the tools TOSTER provides for computing and testing standardized effects assuming they are appropriate.

## Standardized Mean Difference Functions

The `smd_calc` function computes an SMD --- Cohen's $d$, Hedges' $g$ (bias-corrected), or Glass's $\Delta$ --- with confidence intervals and *optional* hypothesis testing (utilizing a z-test: $z = (d - \delta_0)/SE_d$, where $d$ is the sample SMD estimate, $\delta_0$ is the null value being tested, and $SE_d$ is its standard error.). It supports one-sample, paired, and two-sample designs, and can return either an `htest` object or a data frame. The SMD is already reported in `t_TOST` output, but `smd_calc` allows standalone computation and supports equivalence testing directly on the effect size. This may be useful as a complementary function when using `simple_htest` as the primary statistical testing function.

When estimation rather than testing is the goal, `smd_calc` can be called with `alternative = "none"` (the default), which returns only the point estimate and confidence interval without performing a hypothesis test. This is the recommended usage when the primary objective is to report the magnitude of the effect.

```{r}
smd_calc(
  formula = extra ~ group,
  data = sleep,
  paired = TRUE,
  # apply Hedges' correction
  bias_correction = TRUE,
  # apply the repeated measures correction
  rm_correction = TRUE,
  # return an htest object rather than a data frame
  output = "htest"
)
```

When hypothesis testing on the SMD is desired, `smd_calc` supports equivalence and minimal-effect alternatives directly:

```{r}
smd_calc(
  formula = extra ~ group,
  data = sleep,
  paired = TRUE,
  # apply Hedges' correction
  bias_correction = TRUE,
  # apply the repeated measures correction
  rm_correction = TRUE,
  # equivalence hypothesis
  alternative = "equivalence",
  # equivalence bounds on the SMD
  null.value = c(-0.5, 0.5),
  # return an htest object rather than a data frame
  output = "htest"
)
```

However, the parametric z-test used by `smd_calc` relies on distributional assumptions about the sampling distribution of the SMD that may not hold in practice. When a hypothesis test on the SMD is the goal, bootstrapping via `boot_smd_calc` is recommended because it provides more robust confidence intervals and p-values.

```{r}
boot_smd_calc(
  formula = extra ~ group,
  data = sleep,
  paired = TRUE,
  # apply Hedges' correction
  bias_correction = TRUE,
  # apply the repeated measures correction
  rm_correction = TRUE,
  # equivalence hypothesis
  alternative = "equivalence",
  # equivalence bounds on the SMD
  null.value = c(-0.5, 0.5)
)
```


### Checking Assumptions

The SMD functions inherit the assumptions of the underlying t-test: independent observations, approximately normal residuals or differences, and, for two-sample designs, comparable variances across groups. Visual diagnostics from the t-test discussion above apply directly. Two considerations are specific to standardized effect sizes, however. First, outliers and skew affect both the numerator (mean difference) and the denominator (standard deviation) of the SMD, which can amplify their influence on the estimate; the `tr` trimmed-means argument available in the `boot_smd_calc` function is particularly worth considering when the raw data show extreme values. Second, the asymptotic z-test in `smd_calc` relies on an approximation to the sampling distribution of the SMD that can be very inaccurate in small samples, which is the primary motivation for preferring `boot_smd_calc` when hypothesis testing is the goal.

## Equivalence Tests for Correlations

Researchers sometimes need to test that a correlation is practically zero. The `z_cor_test` function provides a asymptotic approach using the Fisher z-transformation. However, this is only an approximate method and is generally *not recommended* as the primary inferential tool for equivalence testing on correlations.

```{r}
z_cor_test(
  x = bugs$LDLF,
  y = bugs$LDHF,
  # type of correlation coefficient
  method = "pearson",
  # two-sided nil hypothesis test
  alternative = "t"
)
```

Because the z-test relies on distributional approximations that can be inaccurate --- particularly with small samples --- bootstrapping via `boot_cor_test` is recommended for more robust inference. The `boot_cor_test` function supports all three major correlation methods (Pearson, Kendall, and Spearman) as well as the Winsorized [@Wilcox_1993] and percentage bend [@Wilcox_1994] correlation coefficients as robust alternatives, and returns standard `htest` objects that integrate with the helper functions described above.

```{r}
res_boot_cor = boot_cor_test(
  x = bugs$LDLF,
  y = bugs$LDHF,
  # type of correlation coefficient
  method = "pearson",
  # equivalence hypothesis
  alternative = "equivalence",
  # equivalence bounds of -0.4 and 0.4
  null = 0.4
)

print(res_boot_cor)
```

```{r, results = "asis"}
cat(describe_htest(res_boot_cor))
```

If only summary statistics are available, an approximate method is provided through `corsum_test`, which uses the Fisher z-transformation to compute confidence intervals and p-values.

### Checking Assumptions

Correlation-based inference in `z_cor_test` and `boot_cor_test` assumes independent paired observations and, for Pearson's coefficient, a roughly linear relationship between variables with approximately symmetric marginal distributions; Spearman and Kendall relax the linearity requirement to monotonicity, making them appropriate when the relationship is non-linear but directionally consistent. The bootstrap version in `boot_cor_test` further relaxes distributional assumptions about the marginals but remains sensitive to observations that exert disproportionate influence on the estimate. Visual diagnostics are the primary tools: scatterplots with a loess smoother to assess linearity or monotonicity, marginal density plots or QQ plots to check distributional shape, and jackknife-after-bootstrap plots to identify individual observations driving bootstrap variability. A worked diagnostic workflow is provided in the online supplement.

## Comparing Standardized Effect Sizes Between Studies

When evaluating whether a replication study produced results consistent with the original, researchers need a way to formally compare effect sizes. The `compare_smd` function tests whether two SMD estimates --- from an original study and a replication, for example --- differ more than would be expected by sampling variability alone.

Consider an original study reporting Cohen's $d_z$ = 0.95 with 25 participants and a replication with $d$ = 0.23 and 50 participants. A researcher might wonder if these these estimates compatible, and it would be prudent to formally test this. Let's say these researchers determine that equivalence between studies would be SMDs within 0.25 of each other. We can then perform a significance and equivalence test between the two original and replication study estimates.

```{r}
# Significance test
compare_smd(
  # SMD and sample size of the original study
  smd1 = 0.95,
  n1 = 25,
  # SMD and sample size of the replication
  smd2 = 0.23,
  n2 = 50,
  # both estimates come from paired designs
  paired = TRUE,
  # nil hypothesis test
  null = 0,
  # significance hypothesis
  alternative = "two.sided"
)

# Equivalence test
compare_smd(
  # SMD and sample size of the original study
  smd1 = 0.95,
  n1 = 25,
  # SMD and sample size of the replication
  smd2 = 0.23,
  n2 = 50,
  # both estimates come from paired designs
  paired = TRUE,
  # equivalence bounds on the difference in SMDs
  null = .25,
  # equivalence hypothesis
  alternative = "equivalence"
)
```

Based on these results, we would reject the null significance hypothesis (conclude the SMDs differ) but fail to reject the equivalence null hypothesis (we cannot conclude the SMDs are practically equivalent). When raw data are available, `boot_compare_smd` provides a bootstrap alternative (see the [package vignettes](https://aaroncaldwell.us/TOSTERpkg/articles/SMD_calcs.html#comparing-smds) for details).

For comparing correlation coefficients between independent studies, the `compare_cor` function provides an analogous test. It uses either the Fisher z-transformation (default) or the Kraatz (Anderson-Hauck) method to compare two correlation coefficients.

```{r}
compare_cor(
  # correlation and degrees of freedom of the first study
  r1 = 0.45,
  df1 = 48,
  # correlation and degrees of freedom of the second study
  r2 = 0.25,
  df2 = 78,
  # equivalence hypothesis
  alternative = "equivalence",
  # equivalence bounds on the difference in correlations
  null = 0.25
)
```

When raw data are available for both studies, `boot_compare_cor` provides a bootstrap alternative that avoids the distributional approximations underlying the Fisher z and Kraatz methods.

The comparison functions `compare_smd` and `compare_cor` inherit the assumptions of the underlying effect size estimates they combine. Because both studies contribute to the comparison, violations in either study propagate into the test. When raw data are available, the bootstrap alternatives (`boot_compare_smd` and `boot_compare_cor`) are preferred because they avoid the Fisher z and related asymptotic approximations, which can be inaccurate particularly when effect sizes are large or sample sizes are modest. When only summary statistics are available, readers should report the diagnostic context of the original and replication studies where possible.

\newpage

# Robust Methods for Equivalence Testing

When t-test assumptions (normality, homoscedasticity) are seriously violated, or when the mean difference is a poor summary of the effect (e.g., with heavy skew or ordinal psychometric scales) alternative approaches are worth considering. I have organized TOSTER into three robust alternative categories: rank-based tests, resampling methods, and log-transformed methods.

## Rank-Based Tests

Rank-based tests operate on the ranks of the data rather than the raw values, which is why they are typically described as non-parametric. It is worth emphasizing that this label applies to the test, not the data. The underlying observations may still come from any distribution, and rank-based methods should not be treated as automatic substitutes for a t-test when normality is in doubt. These tests are not tests of means or medians [@median_test; @karch2021]; they target different estimands (i.e., the quantity a study is attempting to estimate), and the choice between a rank-based test and a t-test should be driven by which estimand is of scientific interest.

### Wilcoxon-Mann-Whitney Tests

The Wilcoxon-Mann-Whitney (WMW) family --- the rank-sum test for two independent groups and the signed-rank test for paired and one-sample designs --- provides non-parametric tests based on ranks. Within the TOST framework, TOSTER performs two directional tests to determine whether the effect lies within (equivalence) or outside (minimal effect) the equivalence bounds. What that effect is depends on an assumption that researchers often do not realize they are making, and there are persistent misconceptions worth addressing directly. The three subsections below separate what the WMW tests evaluate with no additional assumptions, what they evaluate under a location-shift assumption, and what `wilcox_TOST` actually places bounds on.

#### Without a location-shift assumption: a test of stochastic equality

Many researchers believe the WMW tests are tests of medians. They are not. With no assumption beyond independence and exchangeability, the WMW test evaluates stochastic equality which has the null hypothesis that a randomly drawn observation from one group is equally likely to exceed, or to be exceeded by, a randomly drawn observation from the other,

$$
H_0: \space P(X > Y) + \frac{1}{2} \cdot P(X = Y) = 0.5
$$

As @Fay_Malinovsky_2018 demonstrate, the effect size accompanying the test is therefore most naturally expressed as stochastic superiority (or dominance) rather than as a difference in central tendency. Nothing about this null hypothesis refers to a mean or a median.

#### Under a location-shift assumption: a test of central tendency

The location-shift assumption holds that the two distributions are identical in every respect except location. Under this assumption, and only under it, stochastic equality and equality of medians coincide, and the Hodges-Lehmann estimator can be read as an estimate of the median difference. This is the interpretation most readers have in mind when they reach for a WMW test.

When the assumption fails, the two interpretations can come apart dramatically. @median_test demonstrate striking counterexamples: groups can have equal medians yet produce a significant WMW test, very different medians yet a non-significant WMW test, or even significance in the direction opposite the median comparison. Critically, nothing in the test output signals that the assumption has failed --- the p-value and the Hodges-Lehmann estimate are reported the same way either way. The diagnostic described below is therefore not optional if a central-tendency interpretation is intended.

#### What `wilcox_TOST` places bounds on

In TOSTER, `wilcox_TOST` applies the equivalence bounds to the pseudo-median, which corresponds to the mean or median difference only under the location-shift assumption. This has a direct practical consequence for setting bounds: if you are not prepared to assert location shift, the pseudo-median is your estimand, and the smallest effect size of interest must be justified on that scale rather than imported from reasoning about mean differences.

If the hypothesis genuinely concerns means, `perm_t_test` or `boot_t_test` are the more appropriate tools. If it concerns stochastic superiority, `brunner_munzel` provides a clearer framework and a directly interpretable effect size.

```{r}
test1 = wilcox_TOST(
  formula = extra ~ group,
  data = sleep,
  paired = TRUE,
  # equivalence bounds on the pseudo-median
  eqb = .5
)
print(test1)
```

Based on these results, we would conclude there is a significant difference and the results are also not equivalent. But again, the estimate we are testing is the pseudo-median, not the mean or median difference, so this conclusion may not be relevant to the research question at hand.

#### Checking Assumptions

The WMW tests assume independent observations and exchangeability under the null, the latter meaning that group labels carry no information about outcomes if there is no effect. Those two assumptions are all that the stochastic-equality interpretation requires. The two assumptions below determine whether the stronger central-tendency interpretation is available, and each has a straightforward visual diagnostic.

The location-shift assumption --- that the two group distributions differ only in location, not in shape or spread --- is what allows WMW results to be interpreted as a test of the median difference via the Hodges-Lehmann estimator. It is best diagnosed by overlaying the empirical cumulative distribution functions (ECDFs) of the two groups. Under location shift, the curves are horizontal translations of one another, so the horizontal distance between them is constant across the range of the data (Figure \ref{fig:ecdfshift}A). Departures appear as a gap that widens, narrows, or reverses sign, and in the clearest cases as curves that cross (Figure \ref{fig:ecdfshift}B). In panel B, the two groups there differ in both location and spread, so there is no single number that describes "the shift," and the Hodges-Lehmann estimate that `wilcox_TOST` reports does not correspond to a difference in medians.

For the signed-rank variants used by `wilcox_TOST` and `simple_htest`, the difference distribution must additionally be symmetric about the null value for the pseudo-median to coincide with the mean or median difference. Asymmetry is readily checked with a symmetry plot of the paired differences (Figure \ref{fig:symmplot}). As with the location-shift case, visible departures suggest that the test's output --- the pseudo-median and associated p-value --- does not map cleanly onto a hypothesis about central tendency.

```{r }
#| label: ecdfshift
#| echo: false
#| fig.height: 3.8
#| fig.width: 6.5
#| fig.cap: "\\label{fig:ecdfshift}Empirical cumulative distribution functions for two simulated groups, with double-headed arrows marking the horizontal distance between the curves at the 10th, 50th, and 90th percentiles. (A) The location-shift assumption holds: the distributions differ only in location, the arrows are of equal length, and the Hodges-Lehmann estimator can be read as a median difference. (B) The assumption is violated: the groups differ in spread as well as location, so the horizontal gap varies in magnitude and reverses sign across the distribution, and the curves cross. In panel B there is no single shift to estimate, and a central-tendency interpretation of the WMW test is not available."
set.seed(19)
n_ecdf <- 200

ecdf_panel <- function(g1, g2, title, subtitle) {
  df <- data.frame(
    value = c(g1, g2),
    Group = rep(c("Group 1", "Group 2"), c(length(g1), length(g2)))
  )
  probs <- c(0.1, 0.5, 0.9)
  seg <- data.frame(
    p = probs,
    x1 = quantile(g1, probs, type = 7),
    x2 = quantile(g2, probs, type = 7)
  )
  ggplot(df, aes(x = value, colour = Group, linetype = Group)) +
    stat_ecdf(linewidth = 0.7) +
    geom_segment(
      data = seg, inherit.aes = FALSE,
      aes(x = x1, xend = x2, y = p, yend = p),
      arrow = arrow(length = unit(0.05, "in"), ends = "both"),
      linewidth = 0.35, colour = "grey25"
    ) +
    scale_colour_grey(start = 0.1, end = 0.55) +
    coord_cartesian(xlim = c(-7, 8)) +
    labs(x = "Value", y = "Cumulative probability",
         title = title, subtitle = subtitle) +
    theme_minimal(base_size = 10) +
    theme(legend.position = "bottom",
          legend.title = element_blank(),
          plot.title = element_text(size = 10, face = "bold"),
          plot.subtitle = element_text(size = 8))
}

p_shift <- ecdf_panel(
  rnorm(n_ecdf, 0, 1), rnorm(n_ecdf, 1.5, 1),
  "A. Location shift holds",
  "Horizontal gap is constant across the distribution"
)
p_viol <- ecdf_panel(
  rnorm(n_ecdf, 0, 1), rnorm(n_ecdf, 0.5, 2.5),
  "B. Location shift violated",
  "Gap varies and reverses sign; curves cross"
)

p_shift + p_viol + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```


```{r }
#| label: symmplot
#| echo: false
#| fig.height: 5.5
#| fig.width: 5.5
#| fig.cap: "\\label{fig:symmplot}Symmetry plot of the paired differences from the sleep dataset, pairing each observation's distance above the median with the corresponding distance below. Under symmetry, points fall along the 45-degree reference line (dashed); systematic departures — points consistently above or below the line — indicate asymmetry. The signed-rank variant of WMW requires this symmetry for the pseudo-median to coincide with the mean or median of the differences."
# symmetry plot for sleep paired differences
d_sleep <- sleep$extra[sleep$group == 2] - sleep$extra[sleep$group == 1]
m <- median(d_sleep)
sorted <- sort(d_sleep)
upper <- sorted[sorted > m] - m
lower <- m - sorted[sorted < m]
k <- min(length(upper), length(lower))
sp_df <- data.frame(lower = sort(lower)[1:k], upper = sort(upper)[1:k])
mx <- max(sp_df$lower, sp_df$upper)

ggplot(sp_df, aes(x = lower, y = upper)) +
  geom_point(size = 2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  coord_equal(xlim = c(0, mx), ylim = c(0, mx)) +
  labs(x = "Distance below median",
       y = "Distance above median") +
  theme_minimal()
```

\newpage

### Brunner-Munzel Test

The Brunner-Munzel test [@brunner2000] estimates the relative effect $\hat{p} = P(X > Y) + 0.5 \cdot P(X = Y)$, sometimes referred to as the probability of superiority --- the probability that a randomly chosen observation from one group exceeds a randomly chosen observation from the other, with ties split evenly. This estimand is interpretable without distributional assumptions, unlike the WMW test which requires a location-shift assumption to say anything about location differences. @karch2021 argues it should be preferred as the default non-parametric procedure. Its main advantage over WMW is that the null hypothesis of stochastic equality does not require the two distributions to be identical (i.e., the nonparametric Behrens–Fisher problem). WMW's Type I error rate can depart from nominal when the groups are stochastically equal but differ in variance or shape; the Brunner–Munzel test remains valid in that setting and provides a directly interpretable effect size.

Stochastic superiority tests may be particularly well-suited to Likert-type scales which are ubiquitous in psychological research. For a single ordinal item, the spacing between response categories carries no guaranteed meaning, and a mean difference therefore has a strained interpretation. For example, what, concretely, does a 0.4-unit mean difference on a five-point scale communicate? The stochastic superiority estimand sidesteps this entirely. The probability that a randomly selected participant from one group outscores a randomly selected participant from the other is interpretable on its own terms, with no knowledge of the scale required. We are not suggesting that t-tests will *always* mislead in these contexts but it is a concern that should not be quickly dismissed either [@Liddell_Kruschke_2018]. Additionally, even if we assume traditional parametric methods are robust to the ordinal nature of the data, the preserved error rate is not the same as interpretive clarity. For researchers whose primary outcomes live on ordinal scales, the Brunner-Munzel test and its associated effect size --- as well as ordinal models more generally [@Burkner_Vuorre_2019] --- may offer a more natural and defensible default.

The `brunner_munzel` function supports equivalence and minimal effect testing via `alternative`, with bounds on the probability scale via `mu`. Results can be reported on different scales (`"probability"`, `"difference"`, `"logodds"`, or `"odds"`), and three test methods are available: `"t"` (default), `"logit"` (range-preserving CIs), and `"perm"` (studentized permutation; @neubert2007).

```{r}
bm_test = brunner_munzel(
  formula = extra ~ group,
  data = sleep,
  # equivalence hypothesis
  alternative = "equivalence",
  # equivalence bounds 
  # on the stochastic superiority scale
  mu = c(0.3, 0.7)
)
print(bm_test)
```

#### Checking Assumptions

The Brunner-Munzel test has unusually light assumption requirements. It requires independent observations within and between groups, and (for the t-approximation used by default) sample sizes sufficient for the asymptotic approximation to be reasonable. That's essentially it. Unlike WMW, there is no location-shift requirement for the test to be interpretable as a hypothesis about location, and unlike the proportional-odds framing of WMW as a score test from an ordinal regression model, no constant-odds-ratio assumption is needed. The relative effect is defined directly from the two marginal distributions and carries the same interpretation --— the probability that a random observation from one group exceeds one from the other --— regardless of whether the distributions are symmetric, skewed, unimodal, or shaped however they happen to be. Visual inspection of the group distributions remains useful for understanding the data and for choosing a reporting scale, but there is no distributional assumption that these checks are verifying. The studentized permutation variant (`method = "perm"`) is worth considering in small samples, where the t-approximation method may be less trustworthy.


\newpage

## Resampling Methods

When the goal is to test hypotheses about means while relaxing distributional assumptions, resampling methods provide powerful tools. TOSTER implements both bootstrap and permutation approaches through `boot_t_test` and `perm_t_test`, each with distinct inferential logic but a shared motivation to retain the mean as the estimand while obtaining reference distributions empirically --- through resampling --- rather than relying on the normal-theory derivations (i.e., Student's t-test).

Both approaches are natural extensions of the classic t-test. The test statistic itself is unchanged; what changes is how the null distribution is handled. In the conventional t-test, the null distribution of the mean difference is derived analytically under the assumption that the data are drawn from a normal distribution, with equal variances assumed in Student's t-test and unequal variances accommodated in Welch's t-test. The bootstrap approximates the sampling distribution of the same statistic by resampling from the observed data with replacement, whereas the permutation approach constructs an exact or near-exact null distribution by shuffling group labels (two-sample) or flipping signs (one-sample and paired). In both cases, the researcher continues to ask the same question about means; only the calibration of the reference distribution differs. Researchers may find the resampling methods appealing because it follows a familiar testing logic because the estimand, the null hypothesis, and the test statistic remain the same.

Because the reference distribution is generated from the data rather than assumed, distributional requirements are relaxed, though not eliminated. The bootstrap relies on asymptotic arguments and assumes that the empirical distribution is a reasonable stand-in for the population distribution. It performs well with moderate-to-large samples and skewed or heavy-tailed data, but can struggle with very small samples or extreme outliers. The permutation approach is exact under the null hypothesis of exchangeability, meaning group labels carry no information about outcomes. In its basic form this exactness requires identical distributions under the null, which implicitly assumes equal variances. The studentized permutation test [@janssen1997; @chung2013] addresses this limitation by permuting a scale-invariant statistic (essentially the Welch t-statistic), yielding asymptotically valid inference even under heteroscedasticity. @Arboretti_Pesarin_Salmaso_2020 extended this logic to the equivalence testing setting, showing that studentized permutation TOST maintains nominal Type I error rates when variances differ across groups. In short, these methods are assumption-lighter rather than assumption-free, and users should still consider whether these assumptions are reasonable for their design.

A common reflex when normality seems implausible is to switch to the WMW or other rank-based tests. However, these tests answer a different question. As discussed earlier in this manuscript, rank-based procedures test hypotheses about stochastic ordering or pseudo-medians rather than means, and under heteroscedasticity they can reject the null even when the groups have identical means. When the scientific question is genuinely about means --— as it often is in applied research where effect sizes and interventions are naturally expressed on the original scale —-- bootstrap and permutation t-tests should generally be preferred over rank-based methods for handling such violations of assumptions. The rank-based tests remain useful, but as answers to different questions, not as robust substitutes for tests of means.

Choosing between the bootstrap and permutation approaches involves tradeoffs rather than a clear winner in most situations. The studentized permutation test has the strongest theoretical footing for two-sample mean comparisons under heteroscedasticity, and is particularly attractive for equivalence testing [@Arboretti_Pesarin_Salmaso_2020]. The bootstrap offers greater flexibility because it extends naturally to statistics beyond the mean difference, provides readily interpretable confidence intervals via CI inversion [@Thulin_2024], and ensures consistency between reported intervals and p-values. In small samples, permutation approaches tend to have better finite-sample calibration because the null distribution is constructed exactly (or nearly so) from the data at hand, whereas the bootstrap's asymptotic justification can produce liberal tests with very small sample sizes. When outliers are a concern, both methods accept the `tr` argument for trimmed means, which reduces the influence of extreme observations while preserving a location-based interpretation on the original scale. As a rough heuristic, researchers comparing two independent groups with suspected unequal variances may find the studentized permutation test most defensible, while those wanting a single framework that yields coherent intervals and p-values across a range of designs may prefer the bootstrap. Both are substantial improvements over ignoring assumption violations of the t-test, and both return standard `htest` objects that integrate with the TOST framework for equivalence testing.

### Bootstrap t-test

The bootstrap t-test in TOSTER defaults to the studentized bootstrap (`boot_ci = "stud"`) because this matches the original bootstrap t-test proposed by @efron93 (Chapter 16), which is the procedure `boot_t_TOST` implements. In this approach, the t-statistic itself is resampled rather than the raw mean difference, and confidence intervals are constructed from the bootstrap distribution of these pivotal quantities. Other CI methods (percentile, basic, and BCa) remain available via the `boot_ci` argument, but the studentized default preserves consistency with the canonical  @efron93 formulation. It also ensures that the reported p-value and confidence interval are derived from the same bootstrap distribution, so the two will always agree; previous versions of the package used the studentized p-value regardless of the selected CI method, which could produce disagreement between reported intervals and p-values.

```{r}
set.seed(4522)
boot_t_test(
  formula = extra ~ group,
  data = sleep,
  paired = TRUE,
  # equivalence hypothesis
  alternative = "equivalence",
  # equivalence bounds
  mu = c(-0.5, 0.5),
  # number of bootstrap resamples
  R = 999
)
```

The output follows the familiar `htest` format. The reported p-value is the larger of the two one-sided tests against the equivalence bounds, and the confidence interval is the studentized bootstrap CI for the mean difference. Because the p-value is tied to this interval via CI inversion, a researcher reporting both will find the two tell a consistent story. The interval will fall within the equivalence bounds if and only if the equivalence test rejects.
For situations where both a nil hypothesis test and an equivalence test are desired simultaneously with bootstrapping, `boot_t_TOST` returns a `TOSTt` object with bootstrapped SMD estimates [@Kirby2013]:

```{r}
set.seed(891111)
test1 = boot_t_TOST(
  formula = extra ~ group,
  data = sleep,
  paired = TRUE,
  # equivalence bounds
  eqb = .5,
  # number of bootstrap resamples
  R = 999
)
print(test1)
```

The bootstrapped SMDs are a useful feature in their own right. Analytic confidence intervals for standardized effect sizes rely on approximations, which assume normality and can behave poorly with small or skewed samples. Bootstrap intervals for SMDs sidestep these assumptions by resampling the full estimation pipeline, yielding intervals that inherit the robustness properties of the bootstrap itself.

#### Checking Assumptions

The bootstrap t-test makes fewer distributional assumptions than its conventional, analytic, counterpart but is not assumption-free. It requires independent observations and assumes the empirical distribution is a reasonable stand-in for the population distribution. The latter is an assumption that holds better in moderate-to-large samples than in small ones. The studentized bootstrap is reasonably robust to skewness and heavy tails, but remains sensitive to outliers and other high-influence observations, which can be drawn repeatedly across replicates and disproportionately shape the bootstrap distribution. Useful visual checks include the distribution of the data (histograms, boxplots) to identify extreme values, the bootstrap distribution itself to flag problematic skewness or gaps, and the jackknife-after-bootstrap plot to identify individual observations that dominate the bootstrap variability. The `tr` argument for trimmed means provides a straightforward robustness option when outliers are a concern. Diagnostic workflows for all three designs are demonstrated in the online supplement. As with the t-test, these checks reflect assumptions that are relaxed relative to normal-theory inference but not eliminated.

### Permutation t-test

When `R = NULL`, the function enumerates all possible permutations for an exact test; otherwise, it draws R random permutations with the "+1" correction [@phipson2010]. By default the test is studentized (`perm_se = TRUE`), consistent with the recommendations discussed in the methods section above. The `perm_se = FALSE` option is available primarily to match the behavior of other R packages that provide permutation tests that do not studentize by default.

```{r}
set.seed(8812)
perm_t_test(
  formula = extra ~ group,
  data = sleep,
  paired = TRUE,
  # equivalence hypothesis
  alternative = "equivalence",
  # equivalence bounds
  mu = c(-0.5, 0.5),
  # number of random permutations
  R = 999
)
```

Running the bootstrap and permutation equivalence tests on the same paired sleep data using matching equivalence bounds produces broadly similar conclusions, as would be expected. The two approaches can, however, diverge in small samples, under strong heteroscedasticity, or in the presence of influential outliers, which is where the choice between them becomes substantively important. Users interested in robustness to outliers can additionally supply the `tr` argument to either function to use trimmed means, which downweight extreme observations while preserving a location-based interpretation on the original scale.

#### Checking Assumptions

The permutation t-test carries different assumptions depending on design. For the two-sample case, the studentized version is valid under exchangeability of observations across groups under the null and provides asymptotic validity even when variances differ between groups. For paired and one-sample designs, the sign-flip procedure additionally requires that the distribution of differences (or one-sample observations) is symmetric around the null value; asymmetry makes the test approximate rather than exact, though studentization provides some asymptotic protection. Symmetry is a commonly overlooked assumption for paired permutation tests and is easily checked with a symmetry plot of the differences, as demonstrated for the WMW signed-rank case above. For two-sample designs, overlaid empirical cumulative density functions and separate density plots help assess whether group shapes are comparable enough for exchangeability under the null to be credible. Diagnostic workflows for all three designs are demonstrated in the online supplement. As with the bootstrap, these checks reflect assumptions that are relaxed relative to Student's t-test but not eliminated.

\newpage

## Log-Transformed Methods

The natural log transformation stabilizes variance and approximates normality for right-skewed data [@logtest]. An additional advantage is that differences on the log scale back-transform to *ratios*, making interpretation straightforward: if two samples have geometric means of 7 and 10.5, the ratio is 1.5 (i.e., 10.5/7). This property is central to pharmaceutical bioequivalence, where the FDA requires comparisons on the log-transformed scale with equivalence bounds of 0.80 to 1.25 for the ratio of geometric means [@fda; @he2022].

An alternative to `log_TOST` is to log-transform data manually and analyze with `t_TOST`, `simple_htest`, or the resampling methods. Differences on the log scale can then be interpreted as "sympercents" --- symmetric percentage differences on the $100 \cdot \ln(x) - ln(y)$ scale [@cole2000] ---  which are symmetric in a way that ordinary percentage differences are not because percentage changes are calculated from different bases, a 10% loss and a 10% gain do not cancel out, whereas +10 and -10 sympercents are exactly equal in magnitude.^[A concrete illustration that I give my students: if a stock falls 10% on Tuesday and rises 10% from Tuesday on Wednesday, you still have a loss on of your original investment. Sympercents do not have this asymmetry.]

### Log TOST

The `log_TOST` function accepts paired and two-sample comparisons (one-sample tests are not supported since there is no ratio to calculate). Instead of an SMD, a ratio of means is reported [@lajeunesse2015bias]. The default equivalence bounds are set to the FDA standard (`eqb = 1.25`; the reciprocal is taken as the other bound).

```{r, error=FALSE}
log_TOST(mpg ~ am, data = mtcars)
```

\newpage

### Bootstrap Log TOST

The bootstrap version uses the same bootstrap methods as `boot_t_test` on the log-transformed values:

```{r, error=FALSE}
boot_log_TOST(mpg ~ am, data = mtcars, R=999)
```

From this analysis, we would conclude there is a significant effect that is not practically equivalent.

### Checking Assumptions

The log-transformed methods inherit the assumptions of the underlying test applied to the transformed data. The `log_TOST` function carries forward the t-test assumptions, and `boot_log_TOST` carries forward the bootstrap t-test assumptions. What changes is where those assumptions need to hold. Normality, symmetry, and homogeneity of variance should be assessed on the log scale rather than the original scale, since that is where inference takes place. The same visual diagnostics apply just computed on log-transformed data. This points to a broader practical use of log transformation. When the symmetry assumption of the sign-flip permutation test is violated because paired differences are right-skewed on the original scale, the differences of logs are often approximately symmetric, and `perm_t_test` on log-transformed values becomes a reasonable option. In such cases the equivalence bounds should be specified on the log scale and interpreted as log ratios (or sympercents). 

\newpage

# Equivalence Testing with ANOVAs

Many researchers use ANOVA as an omnibus test before inspecting pairwise comparisons, particularly in factorial designs. As @Campbell_2021 note, the lack of a significant ANOVA result does not indicate that a factor has no effect. Equivalence testing for ANOVAs addresses this directly by asking whether the variance attributable to a factor is small enough to be declared negligible.

## Partial Eta-Squared and the Equivalence Hypothesis

The effect size used throughout this section is "partial eta-squared" ($\eta^2_p$),

$$
\eta^2_p = \frac{SS_{\text{effect}}}{SS_{\text{effect}} + SS_{\text{error}}}
$$

the proportion of variance attributable to an effect once the variance attributable to the other effects in the model has been removed. It is not the same quantity as $\eta^2$, which divides by the total sum of squares; the two coincide only in a one-way design. In within-subjects designs, $SS_{\text{error}}$ refers to the error term against which that particular effect is tested, so each effect in a factorial model carries its own denominator.

Because $\eta^2_p$ cannot be negative, there is no lower bound to test against, and the procedure is therefore *not* a two one-sided tests procedure despite the framing used elsewhere in this paper. It is a single one-sided test, which @Campbell_2021 describe as an omnibus non-inferiority test:

$$
H_0: \eta^2_p \geq \Delta \qquad \text{versus} \qquad H_1: \eta^2_p < \Delta
$$

Rejecting $H_0$ supports the claim that the effect in question explains less than $\Delta$ of the variance available to it. There are two limits on that conclusion worth consideration. First, it is a statement about an effect as a whole rather than about any particular contrast within it: an omnibus effect can be negligible while a specific pairwise comparison nested inside it is not, so equivalence at the level of the *F*-test does not license equivalence claims about individual group differences. Second, as with any equivalence test, failing to reject $H_0$ is not evidence that the effect is large and it means only that the data remain compatible with an effect at or beyond the bound.

The test itself is computed from the cumulative distribution function of the non-central *F* distribution. @Campbell_2021 derived the *p*-value for a one-way ANOVA with $J$ groups and $N$ total observations as

$$
p = p_F \left( F; \space J-1, \space N-J, \space \frac{N \cdot \Delta}{1 - \Delta} \right)
$$

and evaluated the Type I error and power of the procedure by simulation. TOSTER generalizes the non-centrality parameter so that the same logic extends to multi-factor, factorial, and within-subjects designs,

$$
\lambda_{eq} = \frac{\Delta}{1 - \Delta} \cdot (df_1 + df_2 + 1), \qquad p_{eq} = p_F(F; \space df_1, \space df_2, \space \lambda_{eq})
$$

where $df_1$ and $df_2$ are the numerator and denominator degrees of freedom for the effect being tested. For a one-way design $df_1 + df_2 + 1 = (J-1) + (N-J) + 1 = N$, so the generalization reduces exactly to the expression given by @Campbell_2021. A fuller derivation is provided in the [package documentation](https://aaroncaldwell.us/TOSTERpkg/articles/the_ftestTOSTER.html).

The quantity $df_1 + df_2 + 1$ is the effective sample size of the error stratum in which an effect is tested, which is not in general the total number of observations; in a mixed design, for example, it recovers the number of subjects for a between-subjects effect. I evaluated the generalization by simulation for one-way within-subjects, factorial between-subjects, and mixed designs, and the equivalence test maintained its nominal Type I error rate at the boundary of the null hypothesis in each. Readers who want the simulation code, the full results, and a longer discussion of interpretation will find them in the online supplement.

The equivalence test inherits the standard ANOVA assumptions, which should be checked on the fitted model itself rather than on the equivalence output. The `performance::check_model()` function provides a comprehensive visual diagnostic suite for most linear model objects in R in a single call; the online supplement demonstrates its use.

One limitation warrants specific mention. For within-subjects factors with more than two levels, `equ_anova` currently uses the uncorrected univariate degrees of freedom and does not apply Greenhouse-Geisser or Huynh-Feldt corrections, even when the fitted object was created with a correction requested. My simulations show that the equivalence test can become substantially anti-conservative when sphericity is badly violated. Sphericity should therefore be assessed on the fitted model, and results treated cautiously when it is in doubt; the supplement quantifies the size of the problem.

## Choosing a Bound on Partial Eta-Squared

Bounds on $\eta^2_p$ are harder to justify than bounds on the original measurement scale (much like the SMD), and the difficulty is worth confronting before running the test.

The first problem is that $\eta^2_p$ is not comparable across designs. Its denominator depends on which other effects appear in the model and, in within-subjects designs, on the error term against which the effect is tested. A partial eta-squared of 0.06 for a one-way between-subjects factor and a partial eta-squared of 0.06 for an interaction in a repeated-measures model are not estimates of the same quantity. A bound therefore cannot simply be borrowed from a published study unless that study used the same design and the same model.

This bites hardest in repeated-measures designs, where subject variance is excluded from the denominator entirely. A bound on partial eta-squared is consequently not a bound on the share of total variance in the data, and the gap between the two widens as the number of levels of the within-subjects factor falls. This is a property of partial eta-squared rather than of the equivalence test --- it applies equally to any partial eta-squared reported from such a design --- but it does mean that a bound reasoned about as "a negligible share of the variance I observe" will not be the bound the test applies. The supplement quantifies the discrepancy.

The second problem is that a proportion of variance has little direct substantive meaning. Unlike a mean difference expressed in the units of the outcome, there is rarely an anchor-based or cost-benefit argument that delivers a number on this scale directly. Researchers who find Cohen's $f$ more intuitive can move between the two with $f^2 = \eta^2_p / (1 - \eta^2_p)$, and the `power_eq_f` function can be used to determine the degrees of freedom required to achieve a given power for a specified bound, or the power attainable for a given design. The choosing-bounds section of the package's F-test vignette discusses the problem in more detail.

The bound of $\Delta = 0.35$ used in both examples below is deliberately large. It was chosen so that the demonstrations produce interpretable output rather than because it represents a defensible threshold, and it should not be read as a default.

## Example with Summary Statistics

Using the `InsectSprays` data and the `equ_ftest` function, which accepts summary *F*-test statistics directly:

```{r warning=FALSE, message=FALSE}
data("InsectSprays")
aovtest = aov(count ~ spray, data = InsectSprays)
anova(aovtest)

```

```{r}
equ_ftest(
  # F statistic and degrees of freedom from the ANOVA table
  Fstat = 34.70228,
  df1 = 5,
  df2 = 66,
  # equivalence bound on partial eta-squared
  eqb = 0.35
)
```

Based on these results, there is a significant effect of "spray" and the differences are *not* statistically equivalent --- we reject the traditional null hypothesis but fail to reject the equivalence null hypothesis.

\newpage

## Example with R ANOVA Objects

The `equ_anova` function accepts objects from `stats::aov`, `car::Anova`, and `afex::aov_car`, avoiding the need to extract summary statistics manually. It returns the familiar ANOVA table with the equivalence test appended:

- `effect`, `df1`, `df2`, `F.value` --- the effect being tested and the components of the traditional *F*-test.
- `p.null` --- the *p*-value for the traditional nil-hypothesis test.
- `pes` --- the estimated partial eta-squared for that effect.
- `eqbound` --- the equivalence bound that was applied.
- `p.equ` --- the *p*-value for the equivalence test.

Because each effect in a within-subjects design is tested against its own error term, each row carries its own $df_2$ and therefore its own denominator for `pes`. Note also that an equivalence test is reported for the intercept; this is rarely a hypothesis of interest and can generally be ignored.

```{r}
# Example using a purely within-subjects design
# (Maxwell & Delaney, 2004, Chapter 12, Table 12.5, p. 578):
library(afex)
data(md_12.1)
aovtest2 = aov_ez(
  # participant identifier
  "id",
  # dependent variable
  "rt",
  # data
  md_12.1,
  # within-subjects factors
  within = c("angle", "noise"),
  anova_table = list(correction = "none", es = "none")
)
equ_anova(aovtest2, eqb = 0.35)
```

\newpage

# Power Analysis and Sample Size Planning

Researchers planning an equivalence study will need to determine the required sample size. Power for equivalence tests behaves differently than for standard tests: larger samples are generally needed, and power depends on both the true effect size and the width of the equivalence bounds.

## Power for TOST t-tests

The `power_t_TOST` function computes exact power for TOST with one-sample, two-sample, and paired t-tests. It can solve for any one of `n`, `power`, or `alpha` given the other parameters are provided. The `delta` argument specifies the true mean difference (set to 0 if the true effect is exactly zero) and `eqb` specifies the equivalence bounds.

For example, to determine the sample size needed to achieve 80% power to detect equivalence within bounds of $\pm$ 0.5 (assuming the true difference is zero and $SD$ = 1):

```{r}
power_t_TOST(
  # true mean difference
  delta = 0,
  # assumed standard deviation
  sd = 1,
  # equivalence bounds
  eqb = 0.5,
  alpha = 0.05,
  # leaving n unspecified solves for the n giving 80% power
  power = 0.8,
  type = "two.sample"
)
```

## Power for Correlations

The `power_z_cor` function provides power analysis for equivalence tests on Pearson correlations using Fisher's z-transformation:

```{r}
power_z_cor(
  # true correlation
  rho = 0,
  # leaving n unspecified solves for the n giving 80% power
  power = 0.8,
  # equivalence bounds of -0.3 and 0.3
  null = 0.3,
  alpha = 0.05,
  # equivalence hypothesis
  alternative = "equivalence"
)
```

## Extensions: PowerTOST and Superpower

For more complex designs --- crossover studies, replicate designs, group-sequential equivalence trials --- the PowerTOST package [@PowerTOST] provides extensive functionality beyond what TOSTER offers. For factorial designs where equivalence contrasts are tested via estimated marginal means, simulation-based power analysis via the Superpower package is recommended [@Lakens_Caldwell_2021]. Both packages complement TOSTER's built-in power tools for common research designs. In some situations, customized simulations may be required to accurately determine power for an equivalence test.

\newpage

# Conclusions

This paper has described the major updates to the TOSTER R package, expanding its scope from the original t-test and correlation equivalence tests to a comprehensive toolkit for frequentist equivalence testing. The update adds a unified t-test interface with minimal effect testing support, a simplified hypothesis testing interface that integrates with the standard R ecosystem, tools for standardized effect sizes and correlations, robust and resampling-based alternatives, ANOVA equivalence testing, and power analysis for sample size planning.

Several limitations should be acknowledged. TOSTER only covers a few standard research designs --- it does not handle many analysis scenarios such as those involving multilevel or mixed-effects models. For these settings, researchers should consider packages like "emmeans" or "marginaleffects", which can accommodate non-zero null hypotheses in more complex modeling frameworks. TOSTER also does not provide Bayesian equivalence testing; researchers interested in Bayesian methods for the null should consult other R packages such as "bayestestR".

I hope that the updated package continues to build upon the original impact of @lakens_ori and makes equivalence testing more accessible to researchers across disciplines. At the very least, I would encourage users to compare results across methods (e.g., Student's t-test vs. permutation t-test vs. bootstrap t-test) to ensure that conclusions do not entirely depend on the chosen analysis. If the interpretation substantially changes depending on the test, this suggests features of the data --- violations of assumptions, outliers, distributional shape --- that warrant further investigation.

\newpage

# Additional Information

The TOSTER package can be installed from CRAN or from GitHub:

```
# Install from CRAN
install.packages("TOSTER")

# Install the development version from GitHub
devtools::install_github("Lakens/TOSTER")
```

## Acknowledgement(s) {-}

I would like to thank everyone from the Lakens laboratory group for their input and suggestions.

## Disclosure statement {-}

The author of this manuscript is the author of the TOSTER package.
Citations of this manuscript will benefit his citation count.

## AI Use Disclosure {-}

During the preparation of the TOSTER package updates and this manuscript, the author used Claude Opus 4.5-4.8 (Anthropic) to assist with editing R code and drafting unit tests. For the manuscript text, Claude Opus was used only for targeted copy-editing — suggesting revisions to specific passages, catching typographical and grammatical errors, and identifying inconsistencies across sections. The manuscript was written by the author, and all AI-assisted output was reviewed, revised, and verified by the author, who takes full responsibility for the content of the package and the manuscript.

## Funding {-}

No funding was provided for this work.

## Notes on contributor(s) {-}

Daniel Lakens provided a review of many of the materials that have been incorporated
into the update of TOSTER, and was the original author of this package.
Without his help and encouragement, the TOSTER package and this update would not exist.

## Nomenclature/Notation {-}

- ANOVA: Analysis of Variance
- Bootstrapping: the use of random sampling with replacement to estimate statistics
- FDA: Food and Drug Administration (United States of America)
- MET: Minimal Effects Test
- ncp: non-centrality parameter
- SESOI: Smallest Effect Size of Interest
- SMD: Standardized Mean Difference (e.g., Cohen's d)
- TOST: Two One-Sided Tests
- WMW: Wilcoxon-Mann-Whitney

## Notes {-}

The R package is also (partially) implemented in [jamovi](https://www.jamovi.org/) as the TOSTER module.

\newpage

# References
