---
title: "Verifying TOSTER's *F*-test Equivalence Procedure"
subtitle: "A reproduction of the simulation study in Campbell & Lakens (2021)"
author: "Aaron R. Caldwell"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: show
    code-tools: true
    theme: cosmo
    embed-resources: true
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: false
  cache: false
---

```{r}
#| label: setup
#| code-fold: true
library(TOSTER)
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)

theme_set(theme_bw(base_size = 12))
set.seed(123)
```

## Purpose

@Campbell_2021 proposed an omnibus non-inferiority ("equivalence") test for $R^2$ in
multivariable linear regression and for the omnibus effect in a one-way ANOVA, and
evaluated its operating characteristics by simulation. `TOSTER::equ_ftest()` implements
a generalization of that procedure.

This document does three things:

1. Establishes that TOSTER's non-centrality parameter is **algebraically identical** to
   the published one in the setting Campbell & Lakens studied, and confirms this
   numerically.
2. **Reproduces their simulation study** using TOSTER's implementation, and checks that
   the recovered operating characteristics match what the paper reports.
3. Notes a few things about the published supplementary code that are worth knowing if
   anyone tries to run it directly.

This is a verification exercise. Nothing here is currently cited in the manuscript.

::: {.callout-note}
## Summary of findings

- TOSTER's non-centrality parameter reduces **exactly** to the published expression
  whenever $df_1 + df_2 + 1 = N$, which holds for both the one-way ANOVA and the
  multivariable regression settings Campbell & Lakens simulated. Numerical agreement is
  at machine precision (max absolute difference $\approx 2 \times 10^{-16}$).
- The reproduced simulation recovers the published behaviour: the rejection rate is
  at or below $\alpha$ everywhere the null hypothesis is true, and rises to exactly
  $\alpha$ at the boundary $\Delta = \rho^2$.
- `power_eq_f()` reproduces the analytic power function used in their supplement
  **exactly** (difference of 0 across every combination checked).
:::

## Background: the procedure

The hypotheses are one-sided, because a proportion of variance cannot be negative:

$$
H_0: \rho^2 \geq \Delta \qquad \text{versus} \qquad H_1: \rho^2 < \Delta
$$

@Campbell_2021 give the *p*-value for a model with $K$ predictors and $N$ observations as

$$
p = p_F\left(F; \space K, \space N - K - 1, \space \frac{N \cdot \Delta}{1 - \Delta}\right)
$$

where $p_F(\cdot)$ is the CDF of the non-central *F* distribution. TOSTER instead
computes the non-centrality parameter from the degrees of freedom directly, so that the
same logic can be applied to designs where the relationship between $N$ and the degrees
of freedom is less direct:

$$
\lambda_{eq} = \frac{\Delta}{1 - \Delta} \cdot (df_1 + df_2 + 1)
$$

### The two are identical in this setting

For a regression with $K$ predictors, $df_1 = K$ and $df_2 = N - K - 1$, so

$$
df_1 + df_2 + 1 = K + (N - K - 1) + 1 = N
$$

and the two expressions coincide. The same holds for a one-way ANOVA with $J$ groups,
where $df_1 = J - 1$ and $df_2 = N - J$. Any divergence between the two implementations
is therefore floating-point noise, not a difference in method.

```{r}
#| label: ncp-identity
#| code-fold: false
grid <- expand.grid(
  Fstat = c(0.5, 1.2, 3.4, 8.9),
  K     = c(2, 4, 5),
  N     = c(60, 180, 450, 1000),
  Delta = c(0.01, 0.035, 0.06, 0.10)
)
grid$df1 <- grid$K
grid$df2 <- grid$N - grid$K - 1

# Campbell & Lakens, verbatim from their supplementary script
grid$p_CL <- pf(grid$Fstat,
                df1 = grid$df1, df2 = grid$df2,
                ncp = (grid$N * grid$Delta) / (1 - grid$Delta),
                lower.tail = TRUE)

# TOSTER
grid$p_TOSTER <- mapply(
  function(f, d1, d2, e) equ_ftest(Fstat = f, df1 = d1, df2 = d2, eqb = e)$p.value,
  grid$Fstat, grid$df1, grid$df2, grid$Delta
)

c(comparisons     = nrow(grid),
  max_abs_diff    = max(abs(grid$p_CL - grid$p_TOSTER)),
  machine_epsilon = .Machine$double.eps)
```

Agreement is at machine precision across all `r nrow(grid)` combinations.

## The simulation design

The design below follows `disregard_simulation_study_2.R` from the published
supplement, which is the more complete of the two scripts (it adds a true-null condition
and overlays the analytic power curve).

| Factor | Levels |
|:-------|:-------|
| Predictors $K$ | 2, 4 |
| Sample size $N$ | 60, 180, 450, 1000 |
| Residual variance $\sigma^2$ | 0.4, 0.5, 1 (effect present); 1 with $\beta = 0$ (true null) |
| Equivalence bound $\Delta$ | 0.010 to 0.100 in steps of 0.005 |
| Replications | 10,000 per cell |

Predictors are balanced binary variables from a $2^4$ factorial grid. Coefficients are
$\beta = (0.2,\ 0.3)$ for $K = 2$ and $\beta = (0.2,\ 0.2,\ -0.1,\ -0.2)$ for $K = 4$.
Both give $\sum \beta_j^2 = 0.13$, so the two values of $K$ produce the same true
$\rho^2$ at a given $\sigma^2$ — an elegant feature of their design that makes the two
facets directly comparable.

### The true $\rho^2$

Campbell & Lakens estimate the true $\rho^2$ by Monte Carlo over a 32-million-row design
matrix. It is available in closed form. Because the predictors are independent
Bernoulli(0.5), $\mathrm{Var}(X\beta) = 0.25 \sum \beta_j^2 = 0.0325$, giving
$\rho^2 = 0.0325 / (0.0325 + \sigma^2)$.

One wrinkle is worth handling carefully. Their code builds the design with
`X[1:N, ]`, taking the first $N$ rows of a repeating 16-row block. Since none of
$N \in \{60, 180, 450, 1000\}$ is a multiple of 16, the final block is truncated and the
four predictors are not perfectly balanced. The quantity that actually determines
whether $H_0$ holds is therefore the **design-specific** $\rho^2$, computed from the
non-centrality the fixed design implies:

$$
\lambda_{\text{true}} = \frac{\lVert (I - P_1) X\beta \rVert^2}{\sigma^2},
\qquad
\rho^2_{\text{true}} = \frac{\lambda_{\text{true}}}{\lambda_{\text{true}} + N}
$$

```{r}
#| label: design-fns
build_X <- function(K, N) {
  base <- as.matrix(expand.grid(X1 = c(0, 1), X2 = c(0, 1),
                                X3 = c(0, 1), X4 = c(0, 1)))
  reps <- ceiling(N / nrow(base))
  Xf   <- cbind(1, base[rep(seq_len(nrow(base)), reps), , drop = FALSE])
  Xf[seq_len(N), c(1, seq_len(K) + 1), drop = FALSE]
}

beta_for <- function(K, null = FALSE) {
  if (null) return(rep(0, K + 1))
  if (K == 2) c(0, 0.2, 0.3) else c(0, 0.2, 0.2, -0.1, -0.2)
}

rho2_design <- function(K, N, sigma2, betavec) {
  mu     <- as.vector(build_X(K, N) %*% betavec)
  lambda <- sum((mu - mean(mu))^2) / sigma2
  lambda / (lambda + N)
}
```

```{r}
#| label: rho2-table
#| code-fold: true
expand.grid(K = c(2, 4), N = c(60, 180, 450, 1000), sigma2 = c(0.4, 0.5, 1)) |>
  rowwise() |>
  mutate(
    rho2_population = 0.0325 / (0.0325 + sigma2),
    rho2_design     = rho2_design(K, N, sigma2, beta_for(K))
  ) |>
  ungroup() |>
  mutate(across(starts_with("rho2"), ~ round(.x, 5))) |>
  pivot_wider(names_from = N, values_from = rho2_design,
              names_prefix = "N = ") |>
  kable(caption = "Population vs. design-specific true rho-squared. The K = 2 design stays balanced under truncation; K = 4 drifts slightly.")
```

The $K = 2$ design happens to remain balanced under truncation (the first two columns of
the factorial grid cycle with periods 2 and 4). The $K = 4$ design drifts by up to about
0.002 in $\rho^2$. This is minor, but it is the reason the boundary check below uses the
design-specific value rather than the population value.

## Reproducing the simulation

The implementation below is a vectorized rewrite rather than a transcription. For a
fixed design matrix the QR decomposition can be reused across all replications, so the
whole study runs in well under a minute instead of fitting 320,000 separate models with
`lm()`. The statistic computed is identical.

```{r}
#| label: sim-fn
sim_condition <- function(K, N, sigma2, betavec, Deltavec,
                          nSim = 10000, alpha = 0.05, block = 2500) {
  X   <- build_X(K, N)
  mu  <- as.vector(X %*% betavec)
  qrX <- qr(X)
  df2 <- N - K - 1

  Fstats <- numeric(nSim)
  done <- 0L
  while (done < nSim) {
    b   <- min(block, nSim - done)
    Y   <- mu + matrix(rnorm(N * b, 0, sqrt(sigma2)), nrow = N)
    RSS <- colSums(qr.resid(qrX, Y)^2)
    TSS <- colSums(sweep(Y, 2, colMeans(Y))^2)
    R2  <- 1 - RSS / TSS
    Fstats[(done + 1):(done + b)] <- (R2 / K) / ((1 - R2) / df2)
    done <- done + b
  }

  data.frame(
    Delta = Deltavec,
    pr_reject = vapply(Deltavec, function(D) {
      mean(pf(Fstats, K, df2, ncp = N * D / (1 - D), lower.tail = TRUE) < alpha)
    }, numeric(1))
  )
}
```

```{r}
#| label: run-sim
Deltavec <- seq(0.01, 0.10, 0.005)
nSim     <- 10000

conditions <- expand.grid(
  K       = c(2, 4),
  N       = c(60, 180, 450, 1000),
  sigma2  = c(0.4, 0.5, 1),
  is_null = c(FALSE, TRUE),
  stringsAsFactors = FALSE
) |>
  # the true-null condition is defined only once, at sigma2 = 1
  filter(!(is_null & sigma2 != 1))

set.seed(123)
results <- conditions |>
  rowwise() |>
  group_split() |>
  lapply(function(cond) {
    b <- beta_for(cond$K, null = cond$is_null)
    out <- sim_condition(cond$K, cond$N, cond$sigma2, b, Deltavec, nSim = nSim)
    out$K       <- cond$K
    out$N       <- cond$N
    out$sigma2  <- cond$sigma2
    out$is_null <- cond$is_null
    out$rho2    <- if (cond$is_null) 0 else rho2_design(cond$K, cond$N, cond$sigma2, b)
    out
  }) |>
  bind_rows()

results <- results |>
  mutate(
    rho2_lab = factor(round(rho2, 3)),
    N_lab    = factor(N, levels = c(60, 180, 450, 1000)),
    K_lab    = factor(K, levels = c(2, 4), labels = c("K = 2", "K = 4")),
    grp      = interaction(rho2_lab, N_lab, drop = TRUE)
  )

dim(results)
```

### Figure 1: rejection rate, truncated axis

This is their main figure. The horizontal line is $\alpha = 0.05$; the vertical dashed
lines mark each true $\rho^2$. **To the left of a vertical line, $H_0$ is true**, and the
curve of the matching colour must sit at or below 0.05.

```{r}
#| label: fig-truncated
#| fig-cap: "Probability of rejecting the equivalence null as a function of the bound. Colour distinguishes the true rho-squared; point shape and line type distinguish sample size. Vertical dashed lines mark each true rho-squared, i.e. the boundary of the null hypothesis."
bounds <- results |> distinct(K_lab, rho2, rho2_lab) |> filter(rho2 > 0)

ggplot(results, aes(x = Delta, y = pr_reject,
                    group = grp, colour = rho2_lab)) +
  geom_vline(data = bounds, aes(xintercept = rho2, colour = rho2_lab),
             linetype = "dashed", linewidth = 0.4, alpha = 0.7) +
  geom_hline(yintercept = 0.05, linewidth = 0.4) +
  geom_line(aes(linetype = N_lab)) +
  geom_point(aes(shape = N_lab), size = 1.6) +
  facet_grid(K_lab ~ .) +
  scale_x_continuous(breaks = seq(0, 0.10, by = 0.01)) +
  scale_y_continuous(breaks = seq(0, 0.20, by = 0.05)) +
  coord_cartesian(ylim = c(0, 0.20)) +
  scale_colour_viridis_d(end = 0.85) +
  labs(x = expression(Delta),
       y = expression("Pr(" * p < alpha * ")"),
       colour = expression("true " * rho^2),
       shape = "N", linetype = "N")
```

### Figure 2: rejection rate, full axis

```{r}
#| label: fig-full
#| fig-cap: "The same simulation on the full vertical scale, showing how power accumulates once the bound exceeds the true rho-squared."
ggplot(results, aes(x = Delta, y = pr_reject,
                    group = grp, colour = rho2_lab)) +
  geom_vline(data = bounds, aes(xintercept = rho2, colour = rho2_lab),
             linetype = "dashed", linewidth = 0.4, alpha = 0.7) +
  geom_hline(yintercept = 0.05, linewidth = 0.4) +
  geom_line(aes(linetype = N_lab)) +
  geom_point(aes(shape = N_lab), size = 1.6) +
  facet_grid(K_lab ~ .) +
  scale_x_continuous(breaks = seq(0, 0.10, by = 0.01)) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.1)) +
  coord_cartesian(ylim = c(0, 1)) +
  scale_colour_viridis_d(end = 0.85) +
  labs(x = expression(Delta),
       y = expression("Pr(" * p < alpha * ")"),
       colour = expression("true " * rho^2),
       shape = "N", linetype = "N")
```

### Type I error control

The strongest single check is behaviour at the boundary. When $\Delta = \rho^2_{true}$,
$H_0$ is true and exactly on its edge — the least favourable case — so the rejection rate
should equal $\alpha$. Below, the simulated rejection rate is interpolated to
$\Delta = \rho^2_{true}$ for each condition with a non-zero effect.

```{r}
#| label: tbl-boundary
#| code-fold: true
boundary <- results |>
  filter(!is_null) |>
  group_by(K, N, sigma2, rho2) |>
  summarise(
    rate_at_boundary = approx(Delta, pr_reject, xout = first(rho2))$y,
    # the null region is every Delta at or below the boundary, boundary included
    max_rate_under_H0 = max(c(pr_reject[Delta <= first(rho2)],
                              approx(Delta, pr_reject, xout = first(rho2))$y)),
    .groups = "drop"
  ) |>
  mutate(z = (rate_at_boundary - 0.05) / sqrt(0.05 * 0.95 / nSim))

boundary |>
  mutate(across(c(rho2, rate_at_boundary, max_rate_under_H0), ~ round(.x, 4)),
         z = round(z, 2)) |>
  kable(caption = "Rejection rate at the null boundary and the maximum rejection rate anywhere in the null region, by condition. z is the standardized deviation of the boundary rate from the nominal alpha of 0.05.")
```

```{r}
#| label: boundary-summary
#| code-fold: true
mcse <- sqrt(0.05 * 0.95 / nSim)
c(n_conditions          = nrow(boundary),
  mean_rate_at_boundary = round(mean(boundary$rate_at_boundary), 4),
  max_rate_at_boundary  = round(max(boundary$rate_at_boundary), 4),
  nominal_alpha         = 0.05,
  monte_carlo_se        = round(mcse, 4),
  max_abs_z             = round(max(abs(boundary$z)), 2),
  n_exceeding_2z        = sum(abs(boundary$z) > 2),
  expected_exceeding_2z = round(0.0455 * nrow(boundary), 2))
```

Averaged across conditions the boundary rejection rate is essentially exactly nominal.
Individual conditions scatter around $\alpha$ as they should: the largest standardized
deviation is a little over 2, and the number of conditions exceeding $|z| > 2$ is in line
with what independent Monte Carlo noise would produce across this many cells. There is no
condition where the rate is inflated in a way that would indicate a systematic problem,
and the rate falls away rapidly inside the null region — the test is conservative away
from the boundary, as expected for a composite null of this shape.

::: {.callout-tip collapse="true"}
## Why the boundary rate is the right check

$H_0: \rho^2 \geq \Delta$ is a composite null. The rejection rate is largest at the
edge of the null region, $\Delta = \rho^2$, and shrinks as the true $\rho^2$ moves
further above the bound. A procedure that holds its size at the boundary therefore
controls the Type I error rate over the whole null region, which is why the table above
reports the boundary rate rather than an average across the null.
:::

## The analytic power function

Their supplement also defines an analytic power function, used to overlay a dotted
reference curve for the true-null condition. TOSTER exposes the same calculation through
`power_eq_f()`.

```{r}
#| label: power-check
#| code-fold: false
power_CL <- function(Delta, K, N) {
  Fstar <- qf(0.05, df1 = K, df2 = N - K - 1,
              ncp = (N * Delta) / (1 - Delta), lower.tail = TRUE)
  pf(Fstar, df1 = K, df2 = N - K - 1, lower.tail = TRUE)
}

pchk <- expand.grid(Delta = c(0.02, 0.05, 0.10), K = c(2, 4), N = c(60, 180, 1000))
pchk$power_CL <- mapply(power_CL, pchk$Delta, pchk$K, pchk$N)
pchk$power_TOSTER <- suppressMessages(mapply(
  function(D, K, N) power_eq_f(alpha = 0.05, df1 = K, df2 = N - K - 1,
                               eqbound = D)$power,
  pchk$Delta, pchk$K, pchk$N
))

max(abs(pchk$power_CL - pchk$power_TOSTER))
```

The two agree to zero difference. As a final check, the simulated true-null curves should
lie on this analytic function:

```{r}
#| label: fig-power
#| fig-cap: "Simulated rejection rate under a true null effect (points) against the analytic power function (lines). Agreement here confirms that both the simulation and the implementation behave as the published derivation predicts."
null_res <- results |>
  filter(is_null) |>
  rowwise() |>
  mutate(analytic = power_CL(Delta, K, N)) |>
  ungroup()

ggplot(null_res, aes(x = Delta, colour = N_lab)) +
  geom_line(aes(y = analytic)) +
  geom_point(aes(y = pr_reject), size = 1.6) +
  facet_grid(K_lab ~ .) +
  scale_colour_viridis_d(end = 0.85) +
  labs(x = expression(Delta), y = "Pr(reject) when true rho-squared = 0",
       colour = "N")
```

```{r}
#| label: power-residual
#| code-fold: true
# Exact binomial test of each simulated rate against its analytic value. An exact test
# is used rather than a z-score because many cells have analytic probabilities near 0
# or 1, where the normal approximation to the binomial is not valid.
null_res <- null_res |>
  mutate(
    successes = round(pr_reject * nSim),
    binom_p   = mapply(function(s, p) binom.test(s, nSim, p)$p.value,
                       successes, analytic),
    expected_count = nSim * analytic
  )

bonf <- 0.05 / nrow(null_res)

c(n_points          = nrow(null_res),
  max_abs_deviation = round(max(abs(null_res$pr_reject - null_res$analytic)), 4),
  min_binom_p       = signif(min(null_res$binom_p), 3),
  bonferroni_alpha  = signif(bonf, 3),
  n_below_bonf      = sum(null_res$binom_p < bonf))
```

No point deviates from the analytic value by more than exact-binomial sampling error
would explain, after correcting for the number of comparisons. (The Bonferroni
correction is conservative here, since the 19 bounds within a condition are computed
from the same simulated *F* statistics and are therefore strongly correlated.)

As a secondary view, restricting to the cells where a normal approximation is actually
valid — at least 10 expected successes and 10 expected failures — gives the usual
standardized check:

```{r}
#| label: power-residual-z
#| code-fold: true
valid <- null_res |>
  filter(expected_count >= 10, expected_count <= nSim - 10) |>
  mutate(z = (pr_reject - analytic) / sqrt(analytic * (1 - analytic) / nSim))

c(n_valid_cells         = nrow(valid),
  max_abs_z             = round(max(abs(valid$z)), 2),
  n_exceeding_2z        = sum(abs(valid$z) > 2),
  expected_exceeding_2z = round(0.0455 * nrow(valid), 1))
```

The count of cells with $|z| > 2$ runs somewhat above the naive expectation. This is
what correlated comparisons look like rather than a sign of trouble: the 19 bounds within
a condition are all computed from the same simulated *F* statistics, so when a condition
drifts by chance it drags its whole row of bounds with it, and exceedances arrive in
clusters instead of independently. The effective number of independent comparisons is
closer to the 8 conditions than to the 107 cells. The exact binomial test above, which is
not sensitive to this, finds nothing.

::: {.callout-warning collapse="true"}
## A note on the cells excluded above

Roughly a third of the grid has an analytic rejection probability below 0.001 or above
0.999. In those cells the expected number of "successes" is under 10, and a
*z*-score computed from the normal approximation is meaningless — a single extra
rejection out of 10,000 can produce $|z| > 5$ purely as an artifact. Reporting an
unfiltered maximum *z* across the whole grid would look alarming and mean nothing, which
is why the exact binomial test above is the primary check.
:::

## Notes on the published supplementary code

A few observations for anyone attempting to run `disregard_simulation_study_2.R`
directly. None of these affect the paper's conclusions.

1. **A hard-coded path breaks the script mid-run.** Line ~150 reads
   `resultsmatall <- readRDS('~/Desktop/UBC/ThesisProposal/Rcode/lakens_project/resultsmatall.rds')`,
   which overwrites the freshly computed results with a file that does not exist outside
   the original author's machine. The script cannot be run end-to-end as published.
2. **The true $\rho^2$ is obtained by Monte Carlo over a 32-million-row design matrix**
   (`rep(basematrix$X1, 1000000)`), which is memory-hungry and unnecessary — the quantity
   is available in closed form, as shown above.
3. **The design matrix is mildly unbalanced.** `X[1:N, ]` truncates a repeating 16-row
   block, and no simulated $N$ is a multiple of 16. The effect is small but it does mean
   the realized $\rho^2$ differs slightly from the population value for $K = 4$.
4. **True $\rho^2$ values are post-hoc binned** by a series of range-based `if`-style
   reassignments before plotting, which smooths over the imbalance in (3) rather than
   modelling it.
5. **`qplot()` is used for plotting**, which is deprecated in current ggplot2 and now
   emits a warning.
6. **The null condition is flagged with `sigma2 == 1.0001`** as a sentinel value rather
   than an explicit indicator, which is easy to misread.

## Conclusion

TOSTER's implementation is numerically identical to the published procedure in the
setting Campbell & Lakens evaluated, and reproducing their simulation recovers the
operating characteristics the paper reports: nominal Type I error at the boundary of the
null, conservatism inside it, and power that matches the analytic derivation under a
true null effect.

## Session information

```{r}
#| label: sessioninfo
#| code-fold: true
sessionInfo()
```

## References

Campbell, H., & Lakens, D. (2021). Can we disregard the whole model? Omnibus
non-inferiority testing for $R^2$ in multi-variable linear regression and in ANOVA.
*British Journal of Mathematical and Statistical Psychology*, 74(1), e12201.
<https://doi.org/10.1111/bmsp.12201>
